---
output: github_document
---
<!-- markdownlint-disable-file -->
<!-- README.md needs to be generated from README.Rmd. Please edit that file -->

# mmrm  <img src="man/figures/logo.svg" align="right" width="175" />

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
\

Mixed models for repeated measures (MMRM) are a popular
choice for analyzing longitudinal continuous outcomes in randomized
clinical trials and beyond; see
[Cnaan, Laird and Slasor (1997)](https://doi.org/10.1002/(SICI)1097-0258(19971030)16:20<2349::AID-SIM667>3.0.CO;2-E)
for a tutorial and
[Mallinckrodt, Lane and Schnell (2008)](https://doi.org/10.1177/009286150804200402)
for a review. This package implements
MMRM based on the marginal linear model without random effects using
Template Model Builder (`TMB`) which enables fast and robust model
fitting. Users can specify a variety of covariance matrices, weight
observations, fit models with restricted or standard maximum
likelihood inference, perform hypothesis testing with Satterthwaite
adjusted degrees of freedom, and extract least square means estimates
by using `emmeans`.

## Main Features

* Responses are assumed normally distributed.
* Covariances:
  * Structures: unstructured, Toeplitz, AR1, compound symmetry, and ante-dependence.
  * Groups: shared covariance structure for all subjects, or group specific covariance structures.
  * Variances: homogeneous or heterogeneous across time points.
* Hypothesis testing:
  * Least square means: `emmeans` package can be used with model outputs to obtain
least square means.
  * Degrees of freedom adjustment: Satterthwaite-adjusted one- and multi-dimensional contrasts.
* Model inference:
  * Supports REML and ML.
  * Supports weights.
  * Automatic changing of optimizer in the case of non-convergence.
  * Manual control of optimization routine.

## Installation

### CRAN

You can install the current stable version from CRAN with:

```{r cran-installation, eval = FALSE}
install.packages("mmrm")
```

### GitHub

You can install the current development version from GitHub with:

```{r gh-installation, eval = FALSE}
if (!require("remotes")) {
  install.packages("remotes")
}
remotes::install_github("openpharma/mmrm")
```

## Getting Started

You can get started by trying out the example:

```{r getting-started}
library(mmrm)
fit <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)
```

This specifies an MMRM with the given covariates and an unstructured covariance
matrix for the timepoints (also called visits in the clinical trial context, here
given by `AVISIT`) within the subjects (here `USUBJID`). While by default this uses
restricted maximum likelihood (REML), it is also possible to use ML, see `?mmrm`.

You can look at the results high-level:

```{r print}
fit
```

The `summary()` method then provides the coefficients table with Satterthwaite
degrees of freedom as well as the covariance matrix estimate:

```{r summary}
summary(fit)
```

## Details

For a more detailed introduction to all of the features of this package, look
at the introduction vignette:

```{r intro-vignette, eval = FALSE}
vignette("introduction")
```

For the available covariance structures, look at the covariance vignette:

```{r cov-vignette, eval = FALSE}
vignette("covariance")
```

In order to understand how `mmrm` is fitting the models, you can read the details
at:

```{r algo-vignette, eval = FALSE}
vignette("algorithm")
```
