---
output: github_document
---
<!-- markdownlint-disable-file -->
<!-- README.md needs to be generated from README.Rmd. Please edit that file -->

# mmrm  <img src="man/figures/logo.svg" align="right" width="175" />

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version-last-release/mmrm](https://www.r-pkg.org/badges/version-last-release/mmrm)
[![CRAN monthly downloads](https://cranlogs.r-pkg.org/badges/mmrm)](https://cranlogs.r-pkg.org/badges/mmrm)
[![CRAN total downloads](https://cranlogs.r-pkg.org/badges/grand-total/mmrm)](https://cranlogs.r-pkg.org/badges/grand-total/mmrm)
<!-- badges: end -->
\

Mixed models for repeated measures (MMRM) are a popular
choice for analyzing longitudinal continuous outcomes in randomized
clinical trials and beyond; see
[Cnaan, Laird and Slasor (1997)](https://doi.org/10.1002/(SICI)1097-0258(19971030)16:20<2349::AID-SIM667>3.0.CO;2-E)
for a tutorial and
[Mallinckrodt, Lane and Schnell (2008)](https://doi.org/10.1177/009286150804200402)
for a review. This package implements
MMRM based on the marginal linear model without random effects using
Template Model Builder (`TMB`) which enables fast and robust model
fitting. Users can specify a variety of covariance matrices, weight
observations, fit models with restricted or standard maximum
likelihood inference, perform hypothesis testing with Satterthwaite
adjusted degrees of freedom, and extract least square means estimates
by using `emmeans`.

**Scope:**

* Continuous responses with normal (but potentially heteroscedastic) residuals.
* Only marginal linear models without individual-level random effects are supported

**Main Features:**

* Flexible covariance specification:
  * [Structures](https://openpharma.github.io/mmrm/main/articles/covariance.html): unstructured, Toeplitz, AR1, compound symmetry, and ante-dependence.
  * Groups: shared covariance structure for all subjects, or group specific covariance structures.
  * Variances: homogeneous or heterogeneous across time points.
* Hypothesis testing:
  * [Least square means](https://openpharma.github.io/mmrm/main/reference/emmeans_support.html): `emmeans` package can be used with model outputs to obtain
least square means and test linear contrasts of model parameters.
  * Degrees of freedom adjustment: Satterthwaite-adjusted one- and multi-dimensional contrasts.
* Model inference:
  * Supports REML and ML.
  * Supports weights.
* Powerful and fast implementation using C++ and automatic differentiation to
obtain precise gradient information for model fitting. See [here](https://openpharma.github.io/mmrm/main/articles/algorithm.html)
for details of the model fitting algorithm used in `mmrm`.


## Installation

**CRAN**

You can install the current stable version from CRAN with:

```{r cran-installation, eval = FALSE}
install.packages("mmrm")
```

**GitHub**

You can install the current development version from GitHub with:

```{r gh-installation, eval = FALSE}
if (!require("remotes")) {
  install.packages("remotes")
}
remotes::install_github("openpharma/mmrm")
```

## Getting Started

See also the [intoductory vignette](https://openpharma.github.io/mmrm/main/articles/introduction.html)
or get started by trying out the example:

```{r getting-started}
library(mmrm)
fit <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)
```

The code specifies an MMRM with the given covariates and an unstructured covariance
matrix for the timepoints (also called visits in the clinical trial context, here
given by `AVISIT`) within the subjects (here `USUBJID`). While by default this uses
restricted maximum likelihood (REML), it is also possible to use ML, see `?mmrm`.

You can look at the results high-level:

```{r print}
fit
```

The `summary()` method then provides the coefficients table with Satterthwaite
degrees of freedom as well as the covariance matrix estimate:

```{r summary}
summary(fit)
```


## Citing `mmrm`

To cite `mmrm` please see [here](https://openpharma.github.io/mmrm/main/authors.html#citation).
