---
output: github_document
---
<!-- markdownlint-disable-file -->
<!-- README.md needs to be generated from README.Rmd. Please edit that file -->

# mmrm  <img src="man/figures/logo.svg" align="right" width="175" />

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![CRAN status](https://www.r-pkg.org/badges/version-last-release/mmrm)](https://www.r-pkg.org/badges/version-last-release/mmrm)
[![CRAN monthly downloads](https://cranlogs.r-pkg.org/badges/mmrm)](https://cranlogs.r-pkg.org/badges/mmrm)
[![CRAN total downloads](https://cranlogs.r-pkg.org/badges/grand-total/mmrm)](https://cranlogs.r-pkg.org/badges/grand-total/mmrm)
[![Code Coverage](https://raw.githubusercontent.com/openpharma/mmrm/_xml_coverage_reports/data/main/badge.svg)](https://raw.githubusercontent.com/openpharma/mmrm/_xml_coverage_reports/data/main/coverage.xml)
<!-- badges: end -->
\

Mixed models for repeated measures (MMRM) are a popular
choice for analyzing longitudinal continuous outcomes in randomized
clinical trials and beyond; see
[Cnaan, Laird and Slasor (1997)](https://doi.org/10.1002/(SICI)1097-0258(19971030)16:20<2349::AID-SIM667>3.0.CO;2-E)
for a tutorial and
[Mallinckrodt, Lane and Schnell (2008)](https://doi.org/10.1177/009286150804200402)
for a review. This package implements
MMRM based on the marginal linear model without random effects using
Template Model Builder (`TMB`) which enables fast and robust model
fitting. Users can specify a variety of covariance matrices, weight
observations, fit models with restricted or standard maximum
likelihood inference, perform hypothesis testing with Satterthwaite
or Kenward-Roger adjustment, and extract least
square means estimates by using `emmeans`.

**Scope:**

* Continuous responses with normal (but potentially heteroscedastic) residuals.
* Marginal linear models (no individual-level random effects).

**Main Features:**

* Flexible covariance specification:
  * [Structures](https://openpharma.github.io/mmrm/main/articles/covariance.html): unstructured, Toeplitz, AR1, compound symmetry, ante-dependence, and spatial exponential.
  * Groups: shared covariance structure for all subjects or group-specific
  covariance estimates.
  * Variances: homogeneous or heterogeneous across time points.
* Hypothesis testing:
  * [Least square means](https://openpharma.github.io/mmrm/main/reference/emmeans_support.html):
    can be obtained with the `emmeans` package
  * One- and multi-dimensional linear contrasts of model parameters can be tested.
  * [Satterthwaite](https://openpharma.github.io/mmrm/main/articles/satterthwaite.html)
    adjusted degrees of freedom.
  * [Kenward-Roger](https://openpharma.github.io/mmrm/main/articles/kenward.html)
    adjusted degrees of freedom and coefficients covariance matrix.
  * [Coefficient Covariance](https://openpharma.github.io/mmrm/main/articles/coef_vcov.html)
* Model inference:
  * Supports REML and ML.
  * Supports weights.
* Fast implementation using C++ and automatic differentiation to
obtain precise gradient information for model fitting. See [here](https://openpharma.github.io/mmrm/main/articles/algorithm.html)
for details of the model fitting algorithm used in `mmrm`.

## Installation

**CRAN**

You can install the current stable version from CRAN with:

```{r cran-installation, eval = FALSE}
install.packages("mmrm")
```

**GitHub**

You can install the current development version from GitHub with:

```{r gh-installation, eval = FALSE}
if (!require("remotes")) {
  install.packages("remotes")
}
remotes::install_github("openpharma/mmrm")
```

## Getting Started

See also the [introductory vignette](https://openpharma.github.io/mmrm/main/articles/introduction.html)
or get started by trying out the example:

```{r getting-started}
library(mmrm)
fit <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)
```

The code specifies an MMRM with the given covariates and an unstructured covariance
matrix for the timepoints (also called visits in the clinical trial context, here
given by `AVISIT`) within the subjects (here `USUBJID`). While by default this uses
restricted maximum likelihood (REML), it is also possible to use ML, see `?mmrm`.

You can look at the results high-level:

```{r print}
fit
```

The `summary()` method then provides the coefficients table with Satterthwaite
degrees of freedom as well as the covariance matrix estimate:

```{r summary}
summary(fit)
```

## Citing `mmrm`

To cite `mmrm` please see [here](https://openpharma.github.io/mmrm/main/authors.html#citation).
