---
output: github_document
---
<!-- markdownlint-disable-file -->
<!-- README.md is generated from README.Rmd. Please edit that file -->

# mmrm  <img src="man/figures/logo.svg" align="right" width="175" />

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
\

The `mmrm` package implements mixed models for repeated measures (MMRM) in R.

## Main Features

* Responses are assumed normally distributed.
* Covariances: 
  * Structures: unstructured, Toeplitz, AR1, compound symmetry, and ante-dependence.
  * Groups: shared covariance structure for all subjects, or shared by group.
  * Supports both time-homogeneous and heterogeneous structures.
* Hypothesis testing:
  * Least square means: `emmeans` package can be used with model outputs to obtain
least square means.
  * Degrees of freedom adjustment: Satterthwaite-adjusted one- and multi-dimensional contrasts.
* Model inference:
  * Supports REML and ML.
  * Supports weights. 
  * Automatic changing of optimizer in the case of non-convergence.
  * Manual control of optimization routine.

## Installation

### GitHub

You can install the current development version from GitHub with:

```{r gh-installation, eval = FALSE}
if (!require("remotes")) {
  install.packages("remotes")
}
remotes::install_github("openpharma/mmrm")
```

## Getting Started

You can get started by trying out the example:

```{r getting-started}
library(mmrm)
fit <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)
```

This specifies an MMRM with the given covariates and an unstructured covariance
matrix for the timepoints (also called visits in the clinical trial context, here
given by `AVISIT`) within the subjects (here `USUBJID`). While by default this uses
restricted maximum likelihood (REML), it is also possible to use ML, see `?mmrm`.

You can look at the results high-level:

```{r print}
fit
```

The `summary()` method then provides the coefficients table with Satterthwaite
degrees of freedom as well as the covariance matrix estimate:

```{r summary}
summary(fit)
```

## Details

For a more detailed introduction to all of the features of this package, look
at the introduction vignette:

```{r intro-vignette, eval = FALSE}
vignette("introduction")
```

For the available covariance structures, look at the covariance vignette:

```{r cov-vignette, eval = FALSE}
vignette("covariance")
```

In order to understand how `mmrm` is fitting the models, you can read the details
at:

```{r algo-vignette, eval = FALSE}
vignette("algorithm")
```
