---
title: "Introduction to `mmrm`"
package: mmrm
author:
  - Daniel Sabanés Bové 
  - Brian M. Lang
output:
  rmarkdown::html_document:
          theme: "spacelab"
          highlight: "kate"
          toc: true
          toc_float: true
vignette: |
  %\VignetteIndexEntry{Introduction to `mmrm`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(mmrm)
```

# Common usage

[A minimal call of mmrm()](https://openpharma.github.io/mmrm/main/reference/mmrm.html),
consisting of only formula and data arguments will produce an object of class 
`mmrm`, `mmrm_fit`, and `mmrm_tmb`. Here we fit a mmrm
model with `us` (unstructured) covariance structure specified, as well as the defaults
of `reml = TRUE` and `optimizer = 'automatic'`.

```{r common-usage-modelcall}
fit <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)
```

Printing the object will show you output which should be familiar to anyone who 
has used any popular modelling functions such as `stats::lm()`, `stats::glm()`,
`glmmTMB::glmmTMB()`, and `lme4::nlmer()`. From this print out we see the function call,
the data used, the covariance structure with number of variance parameters, as well 
as the likelihood method, and model deviance achieved. Additionally the user is 
provided a printout of the estimated coefficients and the model convergence information.

```{r common-usage-print}
print(fit)
```

# Common changes

From the high-level `mmrm()` interface, common changes to the default function 
call can be specified.

## REML or ML

Users can specify if REML should be used (default) or if ML should be used in 
optimization.

```{r common-changes-reml}
fit_ml <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  reml = FALSE
)

print(fit_ml)
```

## Optimizer

Users can specify which optimizer should be used, changing from the default of 
"automatic" selection, which starts with `"L-BFGS-B"` and proceeds through the 
other choices if optimization fails to converge. Other choices are "BFGS", "CG", and
"nlminb". 

```{r common-changes-optim}
fit_opt <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  optimizer = "BFGS"
)

print(fit_opt)
```

## Covariance Structure

Covariance structures supported by the `mmrm` are being continuously developed. 
For a complete list and description please visit the [vignette in the package website](https://openpharma.github.io/mmrm/main/articles/covariance.html). Below
we see the function call for homogeneous compound symmetry (`cs`). 

```{r common-changes-cov}
fit_cs <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + cs(AVISIT | USUBJID),
  data = fev_data,
  reml = FALSE
)

print(fit_cs)
```

# Extraction of model features

Similar to model objects created in other packages, components of `mmrm` and
`mmrm_tmb` objects can be accessed with standard methods.  Additionally, 
`component()` is provided to allow deeper and more precise access for those 
interested in digging through model output.   
Complete documentation of standard model output methods supported for `mmrm_tmb`
objects [can be found at the package website.](https://openpharma.github.io/mmrm/main/reference/mmrm_tmb_methods.html)

## Summary extraction

The `summary` method for `mmrm` objects provides easy access to frequently needed
model components. 

```{r extraction-summary-fit}
fit <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)

fit_summary <- summary(fit)
```

From this summary object, you can easily retrieve the coefficients table.

```{r extraction-summary-fit-coef}
fit_summary$coefficients
```

Other model parameters and metadata available in the summary object is as follows:

```{r extraction-summary-fit-str}
str(fit_summary)
```

## Other components  

Specific model quantities not supported by methods [can be retrieved with the 
`component()` function.](https://openpharma.github.io/mmrm/main/reference/component.html)
The default will output all supported components. 

```{r extraction-summary-component}
str(component(fit))
```

# Lower level functions

## Low-level mmrm

The lower level function which is called by `mmrm()` is `h_mmrm_tmb()`. This 
function is exported and can be used directly. It is similar to `mmrm()` but lacks
some post-processing and support for special d.f. calculations. However, it exposes
an argument for fine control over optimization which may be needed by some
users.

```
h_mmrm_tmb <- function(formula,
  data,
  reml = TRUE,
  control = h_mmrm_tmb_control()
)

h_mmrm_tmb_control <- function(optimizer = stats::nlminb,
  optimizer_args = list(),
  optimizer_control = list(),
  start = NULL,
  accept_singular = TRUE
)
```

# Support for emmeans

This package includes methods that allow `mmrm` objects to be used with the 
`emmeans` package. `emmeans` computes estimated marginal means (also called 
least-square means) for the coefficients of the MMRM.

```{r}
fit <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)

if (require(emmeans)) {
  emmeans(fit, ~ ARMCD | AVISIT)
}
```


# Acknowledgments

The `mmrm` package is based on previous work internal in Roche, namely
the `tern` and `tern.mmrm` packages which were based on `lme4`.
The work done in the `rbmi` package has been important since it used
`glmmTMB` for fitting MMRMs.

We would like to thank Ben Bolker from the `glmmTMB` team for multiple
discussions when we tried to get Satterthwaite degrees of freedom implemented
with `glmmTMB` (see https://github.com/glmmTMB/glmmTMB/blob/satterthwaite_df/glmmTMB/vignettes/satterthwaite_unstructured_example2.Rmd).
Also Ben helped significantly with an example showing how to
use `TMB` for a random effect vector (https://github.com/bbolker/tmb-case-studies/tree/master/vectorMixed).
