---
title: "Introduction to `mmrm`"
package: mmrm
output:
  rmarkdown::html_document:
          theme: "spacelab"
          highlight: "kate"
          toc: true
          toc_float: true
vignette: |
  %\VignetteIndexEntry{Introduction to `mmrm`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(mmrm)
```

# Common usage

A minimal call of [`mmrm()`](https://openpharma.github.io/mmrm/main/reference/mmrm.html),
consisting of only formula and data arguments will produce an object of class
`mmrm`, `mmrm_fit`, and `mmrm_tmb`. Here we fit a mmrm
model with `us` (unstructured) covariance structure specified, as well as the defaults
of `reml = TRUE` and `optimizer = 'automatic'`.

```{r common-usage-modelcall}
fit <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)
```

Printing the object will show you output which should be familiar to anyone who
has used any popular modeling functions such as `stats::lm()`, `stats::glm()`,
`glmmTMB::glmmTMB()`, and `lme4::nlmer()`. From this print out we see the function call,
the data used, the covariance structure with number of variance parameters, as well
as the likelihood method, and model deviance achieved. Additionally the user is
provided a printout of the estimated coefficients and the model convergence information.

```{r common-usage-print}
print(fit)
```

# Common customizations

From the high-level [`mmrm()`](https://openpharma.github.io/mmrm/main/reference/mmrm.html)
interface, common changes to the default function call can be specified.

## REML or ML

Users can specify if REML should be used (default) or if ML should be used in
optimization.

```{r common-changes-reml}
fit_ml <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  reml = FALSE
)

print(fit_ml)
```

## Optimizer

Users can specify which optimizer should be used, changing from the default of
`automatic` selection, which starts with `L-BFGS-B` and proceeds through the
other choices if optimization fails to converge. Other choices are `BFGS`, `CG`, and
`nlminb`.

`L-BFGS-B`, `BFGS` and `CG` are all implemented with `stats::optim()` and the Hessian is not used,
while `nlminb` is using `stats::nlminb() which in turn uses both the gradient and the Hessian for the optimization.

```{r common-changes-optim}
fit_opt <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  optimizer = "BFGS"
)

print(fit_opt)
```

## Covariance Structure

Covariance structures supported by the `mmrm` are being continuously developed.
For a complete list and description please visit the [vignette in the package website](https://openpharma.github.io/mmrm/main/articles/covariance.html). Below
we see the function call for homogeneous compound symmetry (`cs`).

```{r common-changes-cov}
fit_cs <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + cs(AVISIT | USUBJID),
  data = fev_data,
  reml = FALSE
)

print(fit_cs)
```

## Weighting

Users can perform weighted MMRM by specifying a numeric vector `weights` with positive values.

```{r common-changes-weights}
fit_wt <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  weights = fev_data$WEIGHT
)

print(fit_wt)
```

## Grouped Covariance Structure

Grouped covariance structures are supported by the`mmrm` package.
Covariance matrices for each group are identically structured (unstructured, compound symmetry, etc) but the estimates are allowed to vary across groups.
We use the form `cs(time | group / subject)` to specify the group variable.

Here is an example of how we use `ARMCD` as group variable.

```{r group-cov}
fit_cs <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + cs(AVISIT | ARMCD / USUBJID),
  data = fev_data,
  reml = FALSE
)

print(VarCorr(fit_cs))
```

We can see that the estimated covariance matrices are different in different `ARMCD` groups.

# Extraction of model features

Similar to model objects created in other packages, components of `mmrm` and
`mmrm_tmb` objects can be accessed with standard methods.  Additionally,
[`component()`](https://openpharma.github.io/mmrm/main/reference/component.html) is provided to allow deeper and more precise access for those
interested in digging through model output.
Complete documentation of standard model output methods supported for `mmrm_tmb`
objects [can be found at the package website.](https://openpharma.github.io/mmrm/main/reference/mmrm_tmb_methods.html)

## Summary extraction

The `summary` method for `mmrm` objects provides easy access to frequently needed
model components.

```{r extraction-summary-fit}
fit <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)

fit_summary <- summary(fit)
```

From this summary object, you can easily retrieve the coefficients table.

```{r extraction-summary-fit-coef}
fit_summary$coefficients
```

Other model parameters and metadata available in the summary object is as follows:

```{r extraction-summary-fit-str}
str(fit_summary)
```

## Other components

Specific model quantities not supported by methods can be retrieved with the
[`component()`](https://openpharma.github.io/mmrm/main/reference/component.html)
function. The default will output all supported components.

For example, a user may want information about convergence:

```{r extraction-summary-component}
component(fit, name = c("convergence", "evaluations", "conv_message"))
```

or the original low-level call:

```{r}
component(fit, name = "call")
```

the user could also ask for all provided components by not specifying the `name`
argument.

```{r, eval=FALSE}
component(fit)
```

# Lower level functions

## Low-level mmrm

The lower level function which is called by [`mmrm()`](https://openpharma.github.io/mmrm/main/reference/mmrm.html) is [`fit_mmrm()`](https://openpharma.github.io/mmrm/main/reference/fit_mmrm.html). This
function is exported and can be used directly. It is similar to [`mmrm()`](https://openpharma.github.io/mmrm/main/reference/mmrm.html) but lacks
some post-processing and support for Satterthwaite d.f. calculations. However, it exposes
an argument for fine control over optimization which may be needed by some
users.

```{r low-level-hmmrmtmb}
fit_mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  weights = rep(1, nrow(fev_data)),
  reml = TRUE,
  control = mmrm_control()
)
```

For fine control of optimization routine, [`mmrm_control()`](https://openpharma.github.io/mmrm/main/reference/mmrm_control.html)
is provided. This function allows the user to specify optimization routine with
`optimizer`, pass arguments to that optimizer with `optimizer_args`, provide a
list of control parameters with `optimizer_control`, provide a list of starting
parameter values with `start`, and decide the action to be taken when the
defined design matrix is singular with `accept_singular`.

```{r low-level-control, eval=FALSE}
mmrm_control(
  optimizer = stats::nlminb,
  optimizer_args = list(upper = Inf, lower = 0),
  optimizer_control = list(),
  start = c(0, 1, 1, 0, 1, 0),
  accept_singular = FALSE
)
```

# Hypothesis testing

This package supports estimation of one- and multi-dimensional contrasts
(t-test and F-test calculation) with the [`df_1d()`](https://openpharma.github.io/mmrm/main/reference/df_1d.html) and
[`df_md()`](https://openpharma.github.io/mmrm/main/reference/df_md.html) functions.
Both functions utilize Satterthwaite's method for the calculation of test
degrees of freedom.

## One-dimensional contrasts

Compute the test of a one-dimensional (vector) contrast for a `mmrm` object with
Satterthwaite degrees of freedom.

```{r}
fit <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)

contrast <- numeric(length(component(fit, "beta_est")))
contrast[3] <- 1

df_1d(fit, contrast)
```

## Multi-dimensional contrasts

Compute the test of a multi-dimensional (matrix) contrast for a `mmrm` object with
Satterthwaite degrees of freedom.

```{r}
fit <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)

contrast <- matrix(data = 0, nrow = 2, ncol = length(component(fit, "beta_est")))
contrast[1, 2] <- contrast[2, 3] <- 1

df_md(fit, contrast)
```

## Support for emmeans

This package includes methods that allow `mmrm` objects to be used with the
`emmeans` package. `emmeans` computes estimated marginal means (also called
least-square means) for the coefficients of the MMRM.

```{r}
fit <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)

if (require(emmeans)) {
  emmeans(fit, ~ ARMCD | AVISIT)
}
```


# Acknowledgments

The `mmrm` package is based on previous work internal in Roche, namely
the `tern` and `tern.mmrm` packages which were based on `lme4`.
The work done in the `rbmi` package has been important since it used
`glmmTMB` for fitting MMRMs.

We would like to thank Ben Bolker from the `glmmTMB` team for multiple
discussions when we tried to get Satterthwaite degrees of freedom implemented
with `glmmTMB` (see https://github.com/glmmTMB/glmmTMB/blob/satterthwaite_df/glmmTMB/vignettes/satterthwaite_unstructured_example2.Rmd).
Also Ben helped significantly with an example showing how to
use `TMB` for a random effect vector (https://github.com/bbolker/tmb-case-studies/tree/master/vectorMixed).
