---
title: "Introduction to `mmrm`"
package: mmrm
output:
  rmarkdown::html_document:
          theme: "spacelab"
          highlight: "kate"
          toc: true
          toc_float: true
vignette: |
  %\VignetteIndexEntry{Introduction to `mmrm`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(mmrm)
```

# Common usage

A minimal call of [`mmrm()`](https://openpharma.github.io/mmrm/main/reference/mmrm.html),
consisting of only formula and data arguments will produce an object of class
`mmrm`, `mmrm_fit`, and `mmrm_tmb`. Here we fit a mmrm
model with `us` (unstructured) covariance structure specified, as well as the defaults
of `reml = TRUE` and `control = mmrm_control()`.

```{r common-usage-modelcall}
fit <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)
```

Printing the object will show you output which should be familiar to anyone who
has used any popular modeling functions such as `stats::lm()`, `stats::glm()`,
`glmmTMB::glmmTMB()`, and `lme4::nlmer()`. From this print out we see the function call,
the data used, the covariance structure with number of variance parameters, as well
as the likelihood method, and model deviance achieved. Additionally the user is
provided a printout of the estimated coefficients and the model convergence information.

```{r common-usage-print}
print(fit)
```

# Common customizations

From the high-level [`mmrm()`](https://openpharma.github.io/mmrm/main/reference/mmrm.html)
interface, common changes to the default function call can be specified.

## Control Function

For fine control, [`mmrm_control()`](https://openpharma.github.io/mmrm/main/reference/mmrm_control.html)
is provided. This function allows the user to choose the adjustment method for
the degrees of freedom and the coefficients covariance matrix, specify optimization routines,
number of cores to be used on Unix systems for trying several optimizers in parallel,
provide a vector of starting parameter values, decide the action to be taken when the
defined design matrix is singular, not drop unobserved visit levels.
For example:

```{r low-level-control, eval=FALSE}
mmrm_control(
  method = "Kenward-Roger",
  optimizer = c("L-BFGS-B", "BFGS"),
  n_cores = 2,
  start = c(0, 1, 1, 0, 1, 0),
  accept_singular = FALSE,
  drop_visit_levels = FALSE
)
```

Note that this control list can either be passed via the `control` argument
to `mmrm`, or selected controls can be directly specified in the `mmrm` call.
We will see this below.

## REML or ML

Users can specify if REML should be used (default) or if ML should be used in
optimization.

```{r common-changes-reml}
fit_ml <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  reml = FALSE
)
fit_ml
```

## Optimizer

Users can specify which optimizer should be used, changing from the default of
four optimizers, which starts with `L-BFGS-B` and proceeds through the
other choices if optimization fails to converge. Other choices are `BFGS`, `CG`, and
`nlminb`.

`L-BFGS-B`, `BFGS` and `CG` are all implemented with `stats::optim()` and the Hessian
is not used, while `nlminb` is using `stats::nlminb()` which in turn uses both the
gradient and the Hessian (by default but can be switch off) for the optimization.

```{r common-changes-optim}
fit_opt <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  optimizer = "BFGS"
)
fit_opt
```

## Covariance Structure

Covariance structures supported by the `mmrm` are being continuously developed.
For a complete list and description please visit the [covariance vignette](covariance.html).
Below we see the function call for homogeneous compound symmetry (`cs`).

```{r common-changes-cov}
fit_cs <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + cs(AVISIT | USUBJID),
  data = fev_data,
  reml = FALSE
)
fit_cs
```

The time points have to be unique for each subject. That is,
there cannot be time points with multiple observations for any subject.
The rationale is that these observations would need to be correlated, but it
is not possible within the currently implemented covariance structure framework
to do that correctly.

## Weighting

Users can perform weighted MMRM by specifying a numeric vector `weights` with positive values.

```{r common-changes-weights}
fit_wt <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  weights = fev_data$WEIGHT
)
fit_wt
```

## Grouped Covariance Structure

Grouped covariance structures are supported by the`mmrm` package.
Covariance matrices for each group are identically structured (unstructured, compound symmetry, etc) but the estimates are allowed to vary across groups.
We use the form `cs(time | group / subject)` to specify the group variable.

Here is an example of how we use `ARMCD` as group variable.

```{r group-cov}
fit_cs <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + cs(AVISIT | ARMCD / USUBJID),
  data = fev_data,
  reml = FALSE
)
VarCorr(fit_cs)
```

We can see that the estimated covariance matrices are different in different `ARMCD` groups.

## Adjustment Method

Both Satterthwaite and Kenward-Roger adjustment methods are available. The default
is Satterthwaite adjustment of the degrees of freedom. To use Kenward-Roger
adjustment of the degrees of freedom as well as the coefficients covariance
matrix, use the `method` argument:

```{r kr}
fit_kr <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  method = "Kenward-Roger"
)
```

Note that this requires `reml = TRUE`, i.e. Kenward-Roger adjustment is not
possible when using maximum likelihood inference.
While this adjustment choice is not visible in the `print()` result of the fitted model
(because the initial model fit is not affected by the choice of the
adjustment method), looking at the `summary` we see the method and the
correspondingly adjusted standard errors and degrees of freedom:

```{r kr_summary}
summary(fit_kr)
```

For one-dimensional contrasts as in the coefficients table above,
the degrees of freedom are the same for Kenward-Roger and Satterthwaite.
However, Kenward-Roger uses adjusted standard errors, hence the p-values
are different.

Note that if you would like to match SAS results for an unstructured
covariance model, you can use the linear Kenward-Roger approximation:

```{r kr_lin}
fit_kr_lin <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  method = "Kenward-Roger-Linear"
)
```

This is due to the different parametrization of the unstructured covariance
matrix, see the
[Kenward-Roger vignette](kenward.html#parameterization-methods-and-kenward-roger)
for details.

## Keeping Unobserved Visits

Sometimes not all possible time points are observed in a given data set.
When using a structured covariance matrix, e.g. with auto-regressive structure,
then it can be relevant to keep the correct distance between the observed time
points.

Consider the following example where we have deliberately removed the `VIS3`
observations from our initial example data set `fev_data` to obtain `sparse_data`.
We first fit the model where we do not drop the visit level explicitly, using the
`drop_visit_levels = FALSE` choice. Second we fit the model by default without
this option.

```{r sparse}
sparse_data <- fev_data[fev_data$AVISIT != "VIS3", ]
sparse_result <- mmrm(
  FEV1 ~ RACE + ar1(AVISIT | USUBJID),
  data = sparse_data,
  drop_visit_levels = FALSE
)

dropped_result <- mmrm(
  FEV1 ~ RACE + ar1(AVISIT | USUBJID),
  data = sparse_data
)
```

We see that we get a message about the dropped visit level by default.
Now we can compare the estimated correlation matrices:

```{r sparse_cor}
cov2cor(VarCorr(sparse_result))
cov2cor(VarCorr(dropped_result))
```

We see that when using the default, second result, we just drop the `VIS3` from
the covariance matrix. As a consequence, we model the correlation between `VIS2`
and `VIS4` the same as the correlation between `VIS1` and `VIS2`. Hence we get a smaller
correlation estimate here compared to the first result, which includes `VIS3`
explicitly.

# Extraction of model features

Similar to model objects created in other packages, components of `mmrm` and
`mmrm_tmb` objects can be accessed with standard methods.  Additionally,
[`component()`](https://openpharma.github.io/mmrm/main/reference/component.html) is provided to allow deeper and more precise access for those
interested in digging through model output.
Complete documentation of standard model output methods supported for `mmrm_tmb`
objects [can be found at the package website.](https://openpharma.github.io/mmrm/main/reference/mmrm_tmb_methods.html)

## Summary

The `summary` method for `mmrm` objects provides easy access to frequently needed
model components.

```{r extraction-summary-fit}
fit <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)
fit_summary <- summary(fit)
```

From this summary object, you can easily retrieve the coefficients table.

```{r extraction-summary-fit-coef}
fit_summary$coefficients
```

Other model parameters and metadata available in the summary object is as follows:

```{r extraction-summary-fit-str}
str(fit_summary)
```

## Other components

Specific model quantities not supported by methods can be retrieved with the
[`component()`](https://openpharma.github.io/mmrm/main/reference/component.html)
function. The default will output all supported components.

For example, a user may want information about convergence:

```{r extraction-summary-component}
component(fit, name = c("convergence", "evaluations", "conv_message"))
```

or the original low-level call:

```{r}
component(fit, name = "call")
```

the user could also ask for all provided components by not specifying the `name`
argument.

```{r, eval=FALSE}
component(fit)
```

# Lower level functions

## Low-level mmrm

The lower level function which is called by [`mmrm()`](https://openpharma.github.io/mmrm/main/reference/mmrm.html) is [`fit_mmrm()`](https://openpharma.github.io/mmrm/main/reference/fit_mmrm.html). This
function is exported and can be used directly. It is similar to [`mmrm()`](https://openpharma.github.io/mmrm/main/reference/mmrm.html) but lacks
some post-processing and support for Satterthwaite and Kenward-Roger calculations.
It may be useful for other packages that want to fit the model without
the adjustment calculations.

```{r low-level-hmmrmtmb}
fit_mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  weights = rep(1, nrow(fev_data)),
  reml = TRUE,
  control = mmrm_control()
)
```

# Hypothesis testing

This package supports estimation of one- and multi-dimensional contrasts
(t-test and F-test calculation) with the [`df_1d()`](https://openpharma.github.io/mmrm/main/reference/df_1d.html) and
[`df_md()`](https://openpharma.github.io/mmrm/main/reference/df_md.html) functions.
Both functions utilize the chosen adjustment method from the initial `mmrm` call
for the calculation of degrees of freedom and (for Kenward-Roger methods) the
variance estimates for the test-statistics.

## One-dimensional contrasts

Compute the test of a one-dimensional (vector) contrast for a `mmrm` object with
Satterthwaite degrees of freedom.

```{r 1d_satterthwaite}
fit <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)

contrast <- numeric(length(component(fit, "beta_est")))
contrast[3] <- 1

df_1d(fit, contrast)
```

This works similarly when choosing a Kenward-Roger adjustment:

```{r 1d_kr}
fit_kr <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  method = "Kenward-Roger"
)

df_1d(fit_kr, contrast)
```

We see that because this is a one-dimensional contrast, the degrees of freedoms
are identical for Satterthwaite and Kenward-Roger. However, the standard errors
are different and therefore the p-values are different.

## Multi-dimensional contrasts

Compute the test of a multi-dimensional (matrix) contrast for the above defined
`mmrm` object with Satterthwaite degrees of freedom:

```{r md_satterthwaite}
contrast <- matrix(data = 0, nrow = 2, ncol = length(component(fit, "beta_est")))
contrast[1, 2] <- contrast[2, 3] <- 1

df_md(fit, contrast)
```

And for the Kenward-Roger adjustment:

```{r md_kr}
df_md(fit_kr, contrast)
```

We see that for the multi-dimensional contrast we get slightly different
denominator degrees of freedom for the two adjustment methods.

## Support for emmeans

This package includes methods that allow `mmrm` objects to be used with the
`emmeans` package. `emmeans` computes estimated marginal means (also called
least-square means) for the coefficients of the MMRM.

```{r emmeans}
fit <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)

if (require(emmeans)) {
  emmeans(fit, ~ ARMCD | AVISIT)
}
```

# Acknowledgments

The `mmrm` package is based on previous work internal in Roche, namely
the `tern` and `tern.mmrm` packages which were based on `lme4`.
The work done in the `rbmi` package has been important since it used
`glmmTMB` for fitting MMRMs.

We would like to thank Ben Bolker from the `glmmTMB` team for multiple
discussions when we tried to get the Satterthwaite degrees of freedom
implemented with `glmmTMB` (see https://github.com/glmmTMB/glmmTMB/blob/satterthwaite_df/glmmTMB/vignettes/satterthwaite_unstructured_example2.Rmd). Also Ben helped us significantly with
an example showing how to use `TMB` for a random effect vector (https://github.com/bbolker/tmb-case-studies/tree/master/vectorMixed).
