---
title: "Introduction to `mmrm`"
package: mmrm
output:
  rmarkdown::html_document:
          theme: "spacelab"
          highlight: "kate"
          toc: true
          toc_float: true
vignette: |
  %\VignetteIndexEntry{Introduction to `mmrm`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(mmrm)
```

# Common usage

A minimal call of [`mmrm()`](https://openpharma.github.io/mmrm/main/reference/mmrm.html), consisting of only formula and data arguments will produce an object of class `mmrm`, `mmrm_fit`, and `mmrm_tmb`.

Here we fit a mmrm model with `us` (unstructured) covariance structure specified, as well as the defaults of `reml = TRUE` and `control = mmrm_control()`.

```{r, child='.subsections/intro-getting_started.Rmd'}
```

# Common customizations

```{r, child='.subsections/intro-customizations.Rmd'}
```

# Extraction of model features

```{r, child='.subsections/intro-model_features.Rmd'}
```

# Lower level functions

```{r, child='.subsections/intro-lower_level.Rmd'}
```

# Hypothesis testing

```{r, child='.subsections/intro-hypothesis_testing.Rmd'}
```

# Tidymodels

```{r, child='.subsections/intro-tidymodels.Rmd'}
```


# Acknowledgments

```{r, child='.subsections/intro-acknowledgments.Rmd'}
```


# References
