---
title: "Working title for `mmrm` review and comparison"
package: mmrm
bibliography: '`r system.file("REFERENCES.bib", package = "mmrm")`'
output:
  rmarkdown::html_document:
          theme: "spacelab"
          highlight: "kate"
          toc: true
          toc_float: true
vignette: |
  %\VignetteIndexEntry{Working title for `mmrm` review and comparison}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

# Abstract

```{r, child=c('mmrm_review_abstract.Rmd')}
```

# Introduction

Ideas: 
A general, textbook-like, introduction which outlines MMRM and how it works with a 
focus on key references. 
It will provide s plainly written description, and frame mmrm in contrast to 
random slope intercept / glm / ... model. 
Why do we use mmrm rather than other models. 
data types, covariance structures, existing implementations in R and SAS proc mixed
misnomer of "mixed model", since we are actually marginalizing out the random effects

```{r, child=c('mmrm_review_introduction.Rmd')}
```

# Methods

introduction of data sets, need to consider what type of data (simulated or real) 
data problems (low number of observations, varying scales of variables) 
could be quite rigorous.
should at least illustrate that SAS/R are producing similar results on the FEV 
data set. 
illustrate convergence and speed over other non-purpose built models. 

```{r, child=c('mmrm_review_methods.Rmd')}
```

# Results

```{r, child=c('mmrm_review_results.Rmd')}
```

# References
