---
title: "predict intervals"
author: "Liming Li"
output: html_document
editor_options:
  chunk_output_type: console
---

# model

we have the following model

\[
  y = X\beta + \epsilon
\]

and for prediction, we have

\[
  \hat{y_{new}} = X_{new}\hat\beta
\]


where
\[
  \epsilon \sim N(0, \Sigma)
\]
and
\[
  \hat\beta \sim N(\beta, cov(\beta))
\]

# confidence interval

confidence interval is calculated as the variance of $\hat{y}$

\[
  Var(\hat{y_{new}}) = X_{new} cov(\beta) X_{new}^\top
\]

and we can use normal approximation (1.96 +- sd) to obtain the interval.

# prediction interval

for prediction interval, we have

\[
  y_{new} = X_{new}\hat\beta + \epsilon
\]

\[
  Var(y_new|\theta) = X_{new} cov(\beta) X_{new}^\top + \Sigma
\]
where $\Sigma$ and $\beta$ are all function of $\theta$

\[
  E(y_{new}|\theta) = X_{new} \beta
\]

and

\[
  Var(y_{new}) = Var(E(y_{new}|\theta)) + E(Var(y_{new}|\theta))
\]

## Sampling approach

To obtain $Var(y_{new})$, we need to integrate over $\theta$ to obtain the expected variance and vairance of expectation.

However, if we use simulation based approach, we can:

1. Sample $\theta$
2. Obtain $Var(y_{new}|\theta)$ and $E(y_{new}|\theta)$
3. Summarize over $\theta$

to obtain the variance.

# Implementations in `SAS`

```{r}
library(mmrm)
library(sasr)
df2sd(fev_data, "fev_data")
```

```{r}
predict_sas <- function(cov = "un", ddfm = "Satterthewaite") {
  res <- run_sas(
    sprintf("PROC MIXED DATA = fev_data method=reml;
      CLASS RACE(ref = 'Asian') AVISIT(ref = 'VIS4') SEX(ref = 'Male') ARMCD(ref = 'PBO') USUBJID;
      MODEL FEV1 = ARMCD / ddfm=%s solution chisq outpm=predm outp=pred;
      REPEATED AVISIT / subject=USUBJID type=%s r rcorr;
      LSMEANS ARMCD / pdiff=all cl alpha=0.05 slice=AVISIT;
    RUN;", ddfm, cov)
  )
  cat(res$LOG)
  outpm <- sd2df("predm")
  outp <- sd2df("pred")
  res <- data.frame(
    FEV1 = fev_data$FEV1,
    PM = outpm$Pred,
    PMSD = outpm$StdErrPred,
    P = outp$Pred,
    PSD = outp$StdErrPred
  )
  write.csv(res, sprintf("design/predict_interval/%s_%s.csv", cov, ddfm))
}
```

```{r}
predict_sas("un", "satterthwaite")
predict_sas("un", "kenwardroger<linear>")
predict_sas("cs", "satterthwaite")
predict_sas("cs", "kenwardroger")
predict_sas("ar(1)", "satterthwaite")
predict_sas("ar(1)", "kenwardroger")
predict_sas("toep", "satterthwaite")
predict_sas("toep", "kenwardroger")
```
