```{r, include = FALSE}
library(mmrm)
```

From the high-level [`mmrm()`](https://openpharma.github.io/mmrm/main/reference/mmrm.html)
interface, common changes to the default function call can be specified.

## Control Function

For fine control, [`mmrm_control()`](https://openpharma.github.io/mmrm/main/reference/mmrm_control.html)
is provided. This function allows the user to choose the adjustment method for
the degrees of freedom and the coefficients covariance matrix, specify optimization routines,
number of cores to be used on Unix systems for trying several optimizers in parallel,
provide a vector of starting parameter values, decide the action to be taken when the
defined design matrix is singular, not drop unobserved visit levels.
For example:

```{r low-level-control, results = 'hide'}
mmrm_control(
  method = "Kenward-Roger",
  optimizer = c("L-BFGS-B", "BFGS"),
  n_cores = 2,
  start = c(0, 1, 1, 0, 1, 0),
  accept_singular = FALSE,
  drop_visit_levels = FALSE
)
```

Note that this control list can either be passed via the `control` argument
to `mmrm`, or selected controls can be directly specified in the `mmrm` call.
We will see this below.

### Starting Values

The starting values will affect the optimization result. A better starting value
usually can make the optimization more efficient.
In `mmrm` we provide two starting value functions, one is `std_start` and the other is `emp_start`.
`std_start` will try to use the identity matrix as the covariance, however there are convergence problems
for `ar1` and `ar1h` if the identity matrix is provided, thus for these two covariance structures
we use $\rho=0.5$ instead.
`emp_start` will try to use the empirical covariance matrix of the residuals of the ordinary least squares model as the starting value for unstructured covariance structure.
If some timepoints are missing from data, identity matrix will be used for that submatrix.
The correlation between existing and non-existing timepoints are set to 0.

As the starting values will affect the result, please be cautious on choosing the starting values.

#### Example of Default Starting Value Fails

Here we provide an example where the `std_start` fails.
In the following code chunk, we will create a dummy dataset for mmrm analysis.

```{r data_generation}
gen_data <- function(
    n = 100,
    mu = -100 / 52,
    delta = 50 / 52,
    mua = 2000,
    sigmaa = 300,
    sigmab = 60,
    corab = 0.2,
    sigma = 10,
    times = c(0, 2, 6, 12, 24, 36, 52, 70, 88, 104)) {
  nt <- length(times)
  out <- data.frame(
    pts = rep(seq_len(n * 2), each = nt),
    trt = rep(c("Treatment", "Placebo"), rep(n * nt, 2)),
    time = rep(times, n * 2)
  )

  covab <- corab * sigmaa * sigmab # cov between a and b
  cov <- matrix(c(sigmaa^2, covab, covab, sigmab^2), ncol = 2) # Cov matrix for the slope and intercept
  si <- rbind(
    MASS::mvrnorm(n, mu = c(mua, mu + delta), Sigma = cov),
    MASS::mvrnorm(n, mu = c(mua, mu + delta), Sigma = cov)
  )
  idx <- rep(seq_len(n * 2), each = nt)
  out$fev1 <- si[idx, 1] + si[idx, 2] * times + rnorm(n * nt * 2, sd = sigma)
  out$trt <- factor(out$trt)
  out$time <- factor(out$time)
  out$pts <- factor(out$pts)
  return(out)
}
set.seed(123)
out <- gen_data()
```

In the generated data, the variance is not in the same scale across visits.

```{r show_variance}
vapply(split(out$fev1, out$time), sd, FUN.VALUE = 1)
```

Using `emp_start` as the starting value, `mmrm` will converge fast.

```{r mmrm_using_emp_start, eval = !mmrm:::is_r_devel_linux_clang()}
mmrm(fev1 ~ trt * time + us(time | pts), data = out, start = emp_start)
```

However, if we use `std_start`, there will be convergence problems.
We can also force a specific optimization algorithm and add control details, here e.g. choosing `nlminb` with increased maximum number of function evaluations and iterations.

```{r mmrm_using_std_start, eval = FALSE}
mmrm(
  fev1 ~ trt * time + us(time | pts),
  data = out,
  start = std_start,
  optimizer = "nlminb",
  optimizer_control = list(eval.max = 1000, iter.max = 1000)
)
```

## REML or ML

Users can specify if REML should be used (default) or if ML should be used in
optimization.

```{r common-changes-reml}
fit_ml <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  reml = FALSE
)
fit_ml
```

## Optimizer

Users can specify which optimizer should be used, changing from the default of
four optimizers, which starts with `L-BFGS-B` and proceeds through the
other choices if optimization fails to converge. Other choices are `BFGS`, `CG`, `nlminb` and other user-defined custom optimizers.

`L-BFGS-B`, `BFGS` and `CG` are all implemented with `stats::optim()` and the Hessian
is not used, while `nlminb` is using `stats::nlminb()` which in turn uses both the
gradient and the Hessian (by default but can be switch off) for the optimization.

```{r common-changes-optim}
fit_opt <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  optimizer = "BFGS"
)
fit_opt
```

## Covariance Structure

Covariance structures supported by the `mmrm` are being continuously developed.
For a complete list and description please visit the [covariance vignette](covariance.html).
Below we see the function call for homogeneous compound symmetry (`cs`).

```{r common-changes-cov}
fit_cs <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + cs(AVISIT | USUBJID),
  data = fev_data,
  reml = FALSE
)
fit_cs
```

The time points have to be unique for each subject. That is,
there cannot be time points with multiple observations for any subject.
The rationale is that these observations would need to be correlated, but it
is not possible within the currently implemented covariance structure framework
to do that correctly. Moreover, for non-spatial covariance structures, the time
variable must be coded as a factor.

## Weighting

Users can perform weighted MMRM by specifying a numeric vector `weights` with positive values.

```{r common-changes-weights}
fit_wt <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  weights = fev_data$WEIGHT
)
fit_wt
```

## Grouped Covariance Structure

Grouped covariance structures are supported by the`mmrm` package.
Covariance matrices for each group are identically structured (unstructured, compound symmetry, etc) but the estimates are allowed to vary across groups.
We use the form `cs(time | group / subject)` to specify the group variable.

Here is an example of how we use `ARMCD` as group variable.

```{r group-cov}
fit_cs <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + cs(AVISIT | ARMCD / USUBJID),
  data = fev_data,
  reml = FALSE
)
VarCorr(fit_cs)
```

We can see that the estimated covariance matrices are different in different `ARMCD` groups.

## Adjustment Method

In additional to the residual and Between-Within degrees of freedom, both Satterthwaite and Kenward-Roger adjustment methods are available.
The default is Satterthwaite adjustment of the degrees of freedom.
To use e.g. the Kenward-Roger adjustment of the degrees of freedom as well as the coefficients covariance matrix, use the `method` argument:

A list of all allowed `method` is

1. "Kenward-Roger"
1. "Satterthwaite"
1. "Residual"
1. "Between-Within"

```{r kr}
fit_kr <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  method = "Kenward-Roger"
)
```

Note that this requires `reml = TRUE`, i.e. Kenward-Roger adjustment is not
possible when using maximum likelihood inference.
While this adjustment choice is not visible in the `print()` result of the fitted model
(because the initial model fit is not affected by the choice of the
adjustment method), looking at the `summary` we see the method and the
correspondingly adjusted standard errors and degrees of freedom:

```{r kr_summary}
summary(fit_kr)
```

For one-dimensional contrasts as in the coefficients table above,
the degrees of freedom are the same for Kenward-Roger and Satterthwaite.
However, Kenward-Roger uses adjusted standard errors, hence the p-values
are different.

Note that if you would like to match SAS results for an unstructured
covariance model, you can use the linear Kenward-Roger approximation:

```{r kr_lin}
fit_kr_lin <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  method = "Kenward-Roger",
  vcov = "Kenward-Roger-Linear"
)
```

This is due to the different parametrization of the unstructured covariance
matrix, see the
[Kenward-Roger vignette](kenward.html#parameterization-methods-and-kenward-roger)
for details.

## Variance-covariance for Coefficients

There are multiple variance-covariance estimator available for the coefficients, including:

1. "Asymptotic"
1. "Empirical" (Cluster Robust Sandwich)
1. "Empirical-Jackknife"
1. "Empirical-Bias-Reduced"
1. "Kenward-Roger"
1. "Kenward-Roger-Linear"

Please note that, not all combinations of variance-covariance for coefficients and method of
degrees of freedom are possible, e.g. "Kenward-Roger" and "Kenward-Roger-Linear" are available only
when the degrees of freedom method is "Kenward-Roger".

Details can be found in [Coefficients Covariance Matrix Adjustment vignette](coef_vcov.html) and
[Weighted Least Square Empirical Covariance](empirical_wls.html).

An example of using other variance-covariance is:

```{r}
fit_emp <- mmrm(
  formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  method = "Satterthwaite",
  vcov = "Empirical"
)
```

## Keeping Unobserved Visits

Sometimes not all possible time points are observed in a given data set.
When using a structured covariance matrix, e.g. with auto-regressive structure,
then it can be relevant to keep the correct distance between the observed time
points.

Consider the following example where we have deliberately removed the `VIS3`
observations from our initial example data set `fev_data` to obtain `sparse_data`.
We first fit the model where we do not drop the visit level explicitly, using the
`drop_visit_levels = FALSE` choice. Second we fit the model by default without
this option.

```{r sparse}
sparse_data <- fev_data[fev_data$AVISIT != "VIS3", ]
sparse_result <- mmrm(
  FEV1 ~ RACE + ar1(AVISIT | USUBJID),
  data = sparse_data,
  drop_visit_levels = FALSE
)

dropped_result <- mmrm(
  FEV1 ~ RACE + ar1(AVISIT | USUBJID),
  data = sparse_data
)
```

We see that we get a message about the dropped visit level by default.
Now we can compare the estimated correlation matrices:

```{r sparse_cor}
cov2cor(VarCorr(sparse_result))
cov2cor(VarCorr(dropped_result))
```

We see that when using the default, second result, we just drop the `VIS3` from
the covariance matrix. As a consequence, we model the correlation between `VIS2`
and `VIS4` the same as the correlation between `VIS1` and `VIS2`. Hence we get a smaller
correlation estimate here compared to the first result, which includes `VIS3`
explicitly.
