```{r, include = FALSE}
library(mmrm)
```

This package supports estimation of one- and multi-dimensional contrasts
(t-test and F-test calculation) with the [`df_1d()`](https://openpharma.github.io/mmrm/main/reference/df_1d.html) and
[`df_md()`](https://openpharma.github.io/mmrm/main/reference/df_md.html) functions.
Both functions utilize the chosen adjustment method from the initial `mmrm` call
for the calculation of degrees of freedom and (for Kenward-Roger methods) the
variance estimates for the test-statistics.

## One-dimensional contrasts

Compute the test of a one-dimensional (vector) contrast for a `mmrm` object with
Satterthwaite degrees of freedom.

```{r 1d_satterthwaite}
fit <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)

contrast <- numeric(length(component(fit, "beta_est")))
contrast[3] <- 1

df_1d(fit, contrast)
```

This works similarly when choosing a Kenward-Roger adjustment:

```{r 1d_kr}
fit_kr <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data,
  method = "Kenward-Roger"
)

df_1d(fit_kr, contrast)
```

We see that because this is a one-dimensional contrast, the degrees of freedoms
are identical for Satterthwaite and Kenward-Roger. However, the standard errors
are different and therefore the p-values are different.

Additional options for the degrees of freedom `method` are Residual and Between-Within.

## Multi-dimensional contrasts

Compute the test of a multi-dimensional (matrix) contrast for the above defined
`mmrm` object with Satterthwaite degrees of freedom:

```{r md_satterthwaite}
contrast <- matrix(data = 0, nrow = 2, ncol = length(component(fit, "beta_est")))
contrast[1, 2] <- contrast[2, 3] <- 1

df_md(fit, contrast)
```

And for the Kenward-Roger adjustment:

```{r md_kr}
df_md(fit_kr, contrast)
```

We see that for the multi-dimensional contrast we get slightly different
denominator degrees of freedom for the two adjustment methods.

Also the simpler Residual and Between-Within `method` choices can be used of
course together with multidimensional contrasts.

## Support for emmeans

This package includes methods that allow `mmrm` objects to be used with the
`emmeans` package. `emmeans` computes estimated marginal means (also called
least-square means) for the coefficients of the MMRM. For example, in order
to see the least-square means by visit and by treatment arm:

```{r emmeans}
library(emmeans)
lsmeans_by_visit <- emmeans(fit, ~ ARMCD | AVISIT)
lsmeans_by_visit
```

Note that the degrees of freedom choice is inherited here from the initial `mmrm`
fit.
Furthermore, we can also obtain the differences between the treatment arms for each visit
by applying `pairs()` on the object returned by `emmeans()` earlier:

```{r pdiff}
pairs(lsmeans_by_visit, reverse = TRUE)
```

(This is similar like the `pdiff` option in SAS `PROC MIXED`.)
Note that we use here the `reverse` argument to obtain treatment minus
placebo results, instead of placebo minus treatment results.
