```{r, include = FALSE}
library(mmrm)
```

## Tidymodels

`mmrm` is compatible to work in a `tidymodels` workflow. The following is an example of how such a workflow would be constructed.

```{r, warning=FALSE, message=FALSE}
library(tidymodels)
```

### Direct fit

First we define the direct method to fit an mmrm model using the parsnip package functions `linear_reg()` and `set_engine()`.

- `linear_reg()` defines a model that can predict numeric values from predictors using a linear function
- `set_engine()` is used to specify which package or system will be used to fit the model, along with any arguments specific to that software. We can set the method to adjust degrees of freedom directly in the call.

```{r}
model <- linear_reg() |>
  set_engine("mmrm", method = "Satterthwaite") |>
  fit(FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID), fev_data)
```

```{r}
model
```

We can also pass in the full `mmrm_control` object into the `set_engine()` call

```{r}
model_with_control <- linear_reg() |>
  set_engine("mmrm", control = mmrm_control(method = "Satterthwaite")) |>
  fit(FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID), fev_data)
```

### Using mmrm in workflows

We can leverage the `workflows` package in order to fit the same model.

- First we define the specification for linear regression with the mmrm engine.
- Second we define the workflow, by defining the outcome and predictors that will be used in the formula. We then add the model using the formula.
- Lastly, we fit the model

```{r}
mmrm_spec <- linear_reg() |>
  set_engine("mmrm", method = "Satterthwaite")

mmrm_wflow <- workflow() |>
  add_variables(outcomes = FEV1, predictors = c(RACE, ARMCD, AVISIT, USUBJID)) |>
  add_model(mmrm_spec, formula = FEV1 ~ RACE + ARMCD * AVISIT + us(AVISIT | USUBJID))

mmrm_wflow |>
  fit(data = fev_data)
```

We can separate out the data preparation step from the modeling step using the `recipes` package. Here we are converting the `ARMCD` variable into a dummy variable and creating an interaction term with the new dummy variable and each visit.

```{r}
mmrm_recipe <- recipe(FEV1 ~ ., data = fev_data) |>
  step_dummy(ARMCD) |>
  step_interact(terms = ~ starts_with("ARMCD"):AVISIT)
```

Using `prep()` and `juice()` we can see what the transformed data that will be used in the model fit looks like.

```{r}
mmrm_recipe |>
  prep() |>
  juice()
```

We can pass the covariance structure as well in the `set_engine()` definition. This allows for more flexibility on presetting different covariance structures in the pipeline while keeping the data preparation step independent.

```{r}
mmrm_spec_with_cov <- linear_reg() |>
  set_engine(
    "mmrm",
    method = "Satterthwaite",
    covariance = as.cov_struct(~ us(AVISIT | USUBJID))
  )
```

We combine these steps into a workflow:

```{r}
(mmrm_wflow_nocov <- workflow() |>
  add_model(mmrm_spec_with_cov, formula = FEV1 ~ SEX) |>
  add_recipe(mmrm_recipe))
```

Last step is to fit the data with the workflow object

```{r}
(fit_tidy <- fit(mmrm_wflow_nocov, data = fev_data))
```

To retrieve the fit object from within the workflow object run the following

```{r}
fit_tidy |>
  hardhat::extract_fit_engine()
```

## Broom extensions

`mmrm` also contains S3 methods methods for `tidy`, `glance` and `augment` which were
introduced by [`broom`](https://broom.tidymodels.org/). Note that these methods
will work also without loading the `broom` package.
Please see `?mmrm_tidiers` for the detailed documentation.

For example, we can apply the `tidy` method to return a summary table of coefficient estimates:

```{r}
model |> 
  tidy()
```

We can also specify some details to request confidence intervals of specific confidence level:

```{r}
model |> 
  tidy(conf.int = TRUE, conf.level = 0.9)
```

Or we can apply the `glance` method to return a summary table of goodness of fit statistics:

```{r}
model |> 
  glance()
```

Finally, we can use the `augment` method to return a merged tibble of the data, fitted values and residuals:

```{r}
model |> 
  augment()
```

Also here we can specify details for the prediction intervals and type of residuals via the arguments:

```{r}
model |> 
  augment(interval = "confidence", type = "normalized")
```
