---
title: "Simulation Study Resources"
author: "Philippe Boileau and Daniel Sabanes Bove"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, echo=FALSE}
## load required libraries
library(dplyr)
library(stringr)
library(MASS)
library(sasr)

## set global knitr options
knitr::opts_chunk$set(echo = FALSE)
```

Simulation studies are large-scale integration tests used to assess the
correctness of statistical software. This notebook provides information and
guidance on conducting simulation studies in the `mmrm` package. Please review
this resource before performing numerical experiments with `mmrm`.

*This document is a work in progress. Contributions are welcome.*

# The `simChef` Simulation Framework

The [`simChef`](https://yu-group.github.io/simChef/) R package provides a
framework for constructing efficient, high-quality and reproducible simulation
studies. We rely on `simChef` for all numerical experiments. A case study
demonstrating its use is available
[here](https://github.com/PhilBoileau/simChef-case-study). The
`simulations/csh-benchmarks/` directory provides an `mmrm`-specific example.
Please review these resources prior to begining your own numerical experiments.


# Benchmarking `mmrm()`

`mmrm()`'s correctness might be assessed through its comparison to established
software implementing procedures for performing inference about mixed models
with repeated measures. Examples include, but are not limited to, the `nlme`,
`lme4`, and `glmmTMB` R packages and SAS's `PROC MIXED` procedure. Some notes
are provided below to ensure the fair comparison of these software.


## Working with `PROC MIXED`

`mmrm()`'s and SAS's `PROC MIXED` outputs can be directly compared when using
the [`sasr`](https://github.com/insightsengineering/sasr) R package. However,
special consideration is required when handling the repeated measures variable.
If this variable is a factor, then its levels' alphanumeric ordering must
coincide with the expected ordering. An opaque relabeling and reordering of the
repeated measures variable might occur otherwise.

An example is provided in the code blocks below. We begin by defining a
data-generating function:

```{r dgp, echo=TRUE}
#' A Simple RCT Data-Generating Function
#'
#' @description This function simulates a randomized control trial with
#'   participants equally split across the control and treatment arms.
#'   Participants' repeated measures are assumed to be normally distributed with
#'   heterogeneous compound symmetry covariance matrix. Participants' outcomes
#'   are collected at each time point.
#'
#' @param num_part A numeric indicating the total number of participants.
#' @param num_rep_meas A numeric indicating the number of repeated measures.
#' @param fixed_int A numeric corresponding to the fixed intercept effect.
#' @param fixed_trt A numeric corresponding to the fixed treatment effect.
#' @param outcome_vars A numeric vector with length equal to num_rep_meas. It is
#'   the vector of the repeated measures' variances.
#' @param outcome_cor A numeric corresponding to the correlation parameter in
#'   the heterogeneous compound symmetry covariance matrix.
#' @param time_as_integer A logical indicating whether the time point variable
#'   should be kept as an integer. Defaults to FALSE. If set to TRUE, the
#'   pad_zeroes and explicit_time_levels are ignored.
#' @param pad_zeroes A logical indicating whether to pad the time points with a
#'   leading zero. If set to FALSE, the default, then the time variable is an
#'   integer. If set to true, the time variable is transformed into an ordered
#'   factor.
#' @param explicit_time_levels A logical indicating whether the time point
#'   variable's ordering should be explicit. Note that this time point variable
#'   is a factor.
#'
#' @return A list containing the simulated participants' IDs, time point
#'   indicators, outcomes, treatment assignments and baseline covariates.
rct_dgp_fun <- function(
    num_part = 1000,
    num_rep_meas = 10,
    fixed_int = 1,
    fixed_trt = 1,
    outcome_vars = seq_len(num_rep_meas),
    outcome_cor = 0.5,
    time_as_integer = FALSE,
    pad_zeroes = FALSE,
    explicit_time_levels = FALSE) {
  ## form a balanced data.frame
  cov_df <- data.frame(
    participant = seq_len(num_part),
    trt = c(rep(0, round(num_part / 2)), rep(1, num_part - round(num_part / 2)))
  )
  time_point_df <- expand.grid(
    time = seq_len(num_rep_meas),
    participant = seq_len(num_part)
  )
  df <- cov_df %>% dplyr::left_join(time_point_df, by = "participant")

  ## produce the model matrix for the fixed effects
  fixed_model_mat <- model.matrix(~trt, data = df)

  ## define the repeated measures correlation structure, assuming
  ## heterogeneous compound symmetry
  csh_mat <- tcrossprod(sqrt(outcome_vars), sqrt(outcome_vars)) * outcome_cor
  diag(csh_mat) <- outcome_vars

  ## generate the outcomes
  beta <- c(fixed_int, fixed_trt)
  df$y <- fixed_model_mat %*% beta +
    as.vector(t(MASS::mvrnorm(num_part, rep(0, num_rep_meas), csh_mat)))

  ## time point variable transformation
  if (!time_as_integer) {
    if (pad_zeroes) {
      df$time <- stringr::str_pad(df$time, width = 2, side = "left", pad = "0")
    }
    if (explicit_time_levels) {
      df$time <- factor(df$time, levels = as.character(seq_len(num_rep_meas)))
    } else {
      df$time <- as.factor(df$time)
    }
  }

  ## return the generated data
  return(list(
    participant = as.factor(df$participant),
    time = df$time,
    y = df$y,
    trt = df$trt
  ))
}
```

Next, we define a wrapper function for `PROC MIXED` relying on `sasr`:

```{r proc-mixed-fun, echo = TRUE}
#' PROC MIXED Wrapper Function
#'
#' @description This function takes as input the ouput of a data-generating
#'   process function, and fits PROC MIXED repeated measures model on this data.
#'
#' @param participant A factor vector of participant IDs.
#' @param time A factor vector of time points.
#' @param y A numeric vector of outcomes.
#' @param trt A binary vector of treatment group indicators.
#'
#' @return Estimated parameters of the repeated measures' covariance matrix.
#'   These parameters are stored in a data.frame object.
proc_mixed_fun <- function(
    participant,
    time,
    y,
    trt) {
  ## assemble the vectors into a data.frame
  df <- data.frame(
    "participant" = participant,
    "time" = time,
    "y" = y,
    "trt" = trt
  )
  rownames(df) <- NULL

  ## create SAS dataset
  sasr::df2sd(df, "sas_df")

  ## specify the SAS code: only return covariance matrix estimates for
  ## repeated measures
  sas_code <- "ods output CovParms = cov_parms;
    PROC MIXED DATA = sas_df;
      CLASS trt time participant;
      MODEL y = trt;
      REPEATED time / subject=participant type=CSH;
    RUN;"

  ## run the SAS code, and capture the output
  sas_result <- sasr::run_sas(sas_code)
  cov_mat_df <- sasr::sd2df("cov_parms")

  ## prepare the fit object
  return(list(fit = cov_mat_df))
}
```


We then apply this wrapper function to datasets that are identical in all but
their `time` variables. This function returns parameter estimates for the
heterogeneous compound symmetry (CSH) covariance matrix. Eleven parameter
estimates are returned by each procedure: a variance estimate for each time
point, as well as inter-time-point correlation estimate.

In the first dataset, `time` is a factor with levels `1`, `2`, ..., `10`. These
factor levels are not set explicitly. The second dataset is identical to the
first, save that `time`'s factor labels are explicitly provided. The third
dataset's `time` variable is a factor with levels `01`, `02`, ..., `10`. These
factors are not set explicitly. Finally, a fourth dataset is considered in which
`time` is a numeric variable.

```{r run-sasr, echo=TRUE}
## generate datasets with and without time points padded by zeroes
set.seed(123)
rct_data_not_padded <- rct_dgp_fun()
rct_data_not_padded$time %>% head(10)
set.seed(123)
rct_data_not_padded_w_lvls <- rct_dgp_fun(explicit_time_levels = TRUE)
rct_data_not_padded_w_lvls$time %>% head(10)
set.seed(123)
rct_data_padded <- rct_dgp_fun(pad_zeroes = TRUE)
rct_data_padded$time %>% head(10)
set.seed(123)
rct_data_time_integer <- rct_dgp_fun(time_as_integer = TRUE)
rct_data_time_integer$time %>% head(10)

## fit procedure on the non-paded time point data
proc_mixed_fun(
  participant = rct_data_not_padded$participant,
  time = rct_data_not_padded$time,
  y = rct_data_not_padded$y,
  trt = rct_data_not_padded$trt
)

## fit procedure on the non-paded time point data with explicit factor levels
proc_mixed_fun(
  participant = rct_data_not_padded_w_lvls$participant,
  time = rct_data_not_padded_w_lvls$time,
  y = rct_data_not_padded_w_lvls$y,
  trt = rct_data_not_padded_w_lvls$trt
)

## fit procedure to the paded time point data
proc_mixed_fun(
  participant = rct_data_padded$participant,
  time = rct_data_padded$time,
  y = rct_data_padded$y,
  trt = rct_data_padded$trt
)

## fit procedure to the numeric time point data
proc_mixed_fun(
  participant = rct_data_time_integer$participant,
  time = rct_data_time_integer$time,
  y = rct_data_time_integer$y,
  trt = rct_data_time_integer$trt
)
```

When the `time` isn't padded by zeroes---even with explicitly set factor
levels---we find that second and tenth time points' variances are swapped. The
`time` data appears to be loaded as a string into the SAS session and
subsequently sorted. That is, `"1", "2", ..., "10"` is reordered as `"1", "10",
"2", ..., "9"`. That's not the entire story, though, as the names of the
covariance parameters returned by applying the wrapper function to the dataset
with padded `time` aren't padded. As is made clear by the final application,
time points might be fed to SAS as numerics to avoid this issue altogether.

Why is this happening? `sasr` is built on top of
[`SASPy`](https://sassoftware.github.io/saspy/), an interface between SAS and
Python. That means that when an R `data.frame` object is ingested by `sasr`, it
is converted to a [`pandas`
`DataFrame`](https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.html)
prior to being transformed into a SAS data set. Factor variables in the original
`data.frame` are converted to `Categorical` variables in Python, which are then
read recognized as strings by SAS. Hence, if a factor has levels `c(1, 2, ...,
10)`, they are reorderd as `"1", "10", "2", ..., "9"` SAS before being renamed
as `"1", "2", ..., "10"`.


# Session information

```{r session-information}
sessionInfo()
```
