---
title: "Design for empirical covariance"
author: "Liming Li"
output: html_document
editor_options:
  chunk_output_type: console
---

# Implementations in SAS

According to SAS documentation, the robust sandwich estimator is implemented using "empirical" option.
With this option enabled, the method "Satterthwaite" and "Kenward-Roger" are not available.
A simple example in SAS can be

```{r}
library(sasr)

df2sd(fev_data, "fev")

code <- "
ods output diffs = diff;
PROC MIXED DATA = fev empirical cl method=reml;
  CLASS RACE(ref = 'Asian') AVISIT(ref = 'VIS4') SEX(ref = 'Male') ARMCD(ref = 'PBO') USUBJID;
  MODEL FEV1 = ARMCD / solution chisq;
  REPEATED AVISIT / subject=USUBJID type=ar(1) r rcorr;
  LSMEANS ARMCD / pdiff=all cl alpha=0.05 slice=AVISIT;
RUN;"
ret <- run_sas(code)
df <- sd2df("diff")
```

In SAS result, the covariance matrix and degrees of freedom are provided.
The method for degrees of freedom is "between within"

```{r}
cat(ret$LST)
```

# Implementations in `mmrm`

Since using the "empirical" option, we are changing all standard errors and test statistics involving the
fixed effect parameters, `beta_vcov`. So it can be similar to our other methods, "Satterthwaite" and "Kenward-Roger".
So in `mmrm`, the method, stands for both "covariance adjust" and "degrees of freedom calculation".
In SAS, they are separated arguments.

covariance:

1. traditional ${X^\top \Omega^{-1} X}^{-1}$
1. empirical ${X^\top \Omega^{-1} X}^{-1} \sum_{s}{X_i^\top \Omega^{-1} \epsilon_i \epsilon_i^\top \Omega^{-1} X_i} {X^\top \Omega^{-1} X}^{-1}$
1. Kenward-Roger(see KR documentation)

degrees of freedom

1. contain (not implemented)
1. between within (default one)
1. residual (not implemented)
1. Kenward-Roger (this will always use the KR covariance matrix)
1. Satterthwaite (this will always use the traditional covariance matrix)

In our cases, currently we can, still use one option to control them, if we are not going to implement residual/contain covariance matrix.
Of course, we will not have the option to combine the between within degrees of freedom + traditional covariance matrix.

```{r}
library(Matrix)

fit <- mmrm(FEV1 ~ ARMCD + ar1(AVISIT|USUBJID), data = fev_data)

visits <- fit$tmb_data$visits_zero_inds
sel_mat <- function(vis, n_vis) {
  ret <- matrix(0, nrow = length(vis), ncol = n_vis)
  for (i in seq_len(length(vis))) {
    ret[i, vis[i]] <- 1
  }
  return(ret)
}

sel_matrix <- sel_mat(visits + 1L, 4)
sel_matrix_i <- lapply(seq_len(length(fit$tmb_data$subject_zero_inds)), function(x) {
  sel_matrix[(fit$tmb_data$subject_zero_inds[x]+1):(fit$tmb_data$subject_zero_inds[x]+fit$tmb_data$subject_n_visits[x]),, drop = FALSE]
})
v <- lapply(seq_len(length(sel_matrix_i)), function(i) {
  sel_matrix_i[[i]] %*% fit$cov %*% t(sel_matrix_i[[i]])
})
vinv <- lapply(v, solve)

vall <- bdiag(vinv)
x <- fit$tmb_data$x_matrix
eps <- fit$tmb_data$y_vector - fit$tmb_data$x_matrix %*% fit$beta_est
bread <- fit$beta_vcov #solve(t(x) %*% vall %*% x)

meat <- lapply(seq_len(length(sel_matrix_i)), function(i) {
  ii <- (fit$tmb_data$subject_zero_inds[i]+1):(fit$tmb_data$subject_zero_inds[i]+fit$tmb_data$subject_n_visits[i])
  t(x[ii, , drop = FALSE]) %*% vinv[[i]] %*% eps[ii, , drop = FALSE] %*% t(eps[ii, drop = FALSE]) %*% vinv[[i]] %*% x[ii, , drop = FALSE]
})
meat_sum <- Reduce(`+`, meat)

empirical <- bread %*% meat_sum %*% bread
```

To achieve this, the following is proposed

```{r}
fit <- mmrm(FEV1 ~ ARMCD + ar1(AVISIT|USUBJID), data = fev_data, method = "empirical")
```

with this, the covariance matrix are computed.

# Implementations in `nlme` and `clubSandwich`

To run a similar model, we can use `gls` to create a AR1 mixed model.
Then we can use `vcovCR` to obtain the empirical covariance matrix.
Please note that `type = "CR0"` is used to match SAS results(SAS do not do small sample size correction).
```{r}
library(nlme)
m <- gls(
  FEV1 ~ ARMCD,
  correlation = corAR1(form = ~1 | USUBJID),
  data = fev_data, 
  na.action = na.omit, 
  method = "REML"
)
clubSandwich::vcovCR(m, type = "CR0")
```

