---
title: "Hypothese Testing (Three types of ANOVA) for MMRM"
output: html_document
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---


# Background

This [blog](https://www.r-bloggers.com/2011/03/anova-%E2%80%93-type-iiiiii-ss-explained/) has a nice summary of it.
The original [post](http://goanna.cs.rmit.edu.au/fscholer/anova.php) is not accessible now.

And we can also refer to this [page](https://agstats.io/tutorials/sas-proc-mixed)

Some available code from [glmmTMB](https://github.com/glmmTMB/glmmTMB/blob/master/glmmTMB/R/Anova.R)

SAS also has some paper on this topic, but from a different angle.
It uses estimable functions to construct the three types of hypothesis.

See paper https://doi.org/10.1080/03610928008827869 and https://support.sas.com/documentation/onlinedoc/stat/131/introglmest.pdf.

## SAS results

```{r}
sas_code <- "
ods output Tests1 = Tests1 Tests2 = Tests2 Tests3 = Tests3;

PROC MIXED DATA = fev cl method=reml;
  CLASS AVISIT(ref = 'VIS4') ARMCD(ref = 'PBO') USUBJID;
  MODEL FEV1 = ARMCD*AVISIT ARMCD AVISIT  / ddfm=satterthwaite htype=1,2,3 e3;
  REPEATED AVISIT / subject=USUBJID type=ar(1);
RUN;
"

library(sasr)
library(mmrm)
library(car)
df2sd(fev_data, "fev")
res <- run_sas(sas_code)
df3 <- sd2df("tests3")
df2 <- sd2df("tests2")
```

Please note that type 1 result is dependent on the order of testing so it is not included here for simplicity.
Type 2 and 3 tests are independent of the order of model parameter provided to the procedure.

Type 2

```{r}
df2
```

Type 3

```{r}
df3
```


## emmeans joint_tests

```{r}
library(mmrm)
fit <- mmrm(FEV1 ~ ARMCD + AVISIT + ARMCD * AVISIT + ar1(AVISIT | USUBJID), data = fev_data)
emmeans::joint_tests(fit)
```


however, the result do not agree well between SAS and R.

Using "satterthwaite df" the F statistic is close, the ddf is not.

## Finding suitable contrasts

In `emmeans` the contrasts can be obtained for Type 3, and
use these contrasts to send to `df_md` we get exactly the same results with SAS.

```{r}
args <- emmeans:::.zap.args(object = fit, cov.reduce = emmeans:::meanint, omit = "submodel")
object <- do.call(emmeans:::ref_grid, args)
by <- NULL
facs <- setdiff(names(object@levels), c(by, "1"))
emm <- emmeans::emmeans(fit, facs[2], by = by)
emmgrid <- emmeans::contrast(emm, interaction = "consec", by = union(by, NULL))
tst <- emmeans::test(emmgrid, by = by, joint = TRUE, status = TRUE)
ef <- attr(tst, "est.fcns")
df_md(fit, ef$all) # different from joint_tests(fit)
```

This is the contrast used. Please note that this contrast is normalized, however the result is still the same.

```
$ARMCD
     [,1]       [,2] [,3] [,4] [,5]       [,6]       [,7]       [,8]
[1,]    0 -0.9176629    0    0    0 -0.2294157 -0.2294157 -0.2294157

$`ARMCD:AVISIT`
     [,1] [,2] [,3] [,4] [,5] [,6] [,7] [,8]
[1,]    0    0    0    0    0   -1    0    0
[2,]    0    0    0    0    0    0   -1    0
[3,]    0    0    0    0    0    0    0   -1

$AVISIT
     [,1] [,2]       [,3]       [,4]       [,5]       [,6]       [,7]
[1,]    0    0 -0.8944272  0.0000000  0.0000000 -0.4472136  0.0000000
[2,]    0    0  0.0000000 -0.8944272  0.0000000  0.0000000 -0.4472136
[3,]    0    0  0.0000000  0.0000000 -0.8944272  0.0000000  0.0000000
  [,8]
[1,]  0.0000000
[2,]  0.0000000
[3,] -0.4472136
```

### Initial thoughts

Using doolittle method to obtain the contrasts seems plausible? but this is initial.

```{r}
library(checkmate)
doolittle <- function(m) {
  assert_matrix(m)
  for (i in seq_len(nrow(m))) {
    for (j in seq_len(i - 1)) {
      m[i, ] <- m[i, ] - m[i, j] * m[j, ]
    }
    if (m[i, i] == 0) {
      m[i, ] <- 0
    } else {
      m[i, ] <- m[i, ] / m[i, i]
    }
  }
  return(m)
}

normalize_mat <- function(m) {
  m / rowSums(m^2)
}

mx <- model.matrix(fit)
dl <- doolittle(t(mx) %*% mx)
df_md(fit, dl[6:8, ])
df_md(fit, normalize_mat(dl[6:8, ]))
df_md(fit, dl[2, ]) # d.f. do not match, stat matches a lot ()
```

Note: the doolittle method is used to create type 1 contrasts. Not really relavent here.


# Type II Estimable Functions

According to the SAS/STAT® 13.1 User’s Guide, type II estimable can be achieved through

For fixed effect F1:

1. All columns of L associated with effects not containing F1 (except F1) are zero.
2. The submatrix of L associated with effect F1 is $(X_1^\top M X_1)^{-}X_1^\top M X_1$
3. Each of the remaining submatrices of L associated with an effect F2 that contains F1 is $(X_1^\top M X_1)^{-}X_1^\top M X_2$

Note: $X^{-}$ is the generalized g2-inverse of $X$ if $X$ is a square matrix. And $(X^\top X)^{-}X^\top X$ is used
in `PROC GLM` to represent `L` symbolically.
And
$X_0$ : columns of $X$ whose associated effects do not contain F1
$X_1$ : columns of $X$ associated with F1
$X_2$ : columns of $X$ associated with F2 that contains F1
$M = I - X_0(X_0^\top X_0)^{-}X_0^\top$.

for the following mmrm model
```{r}
fit <- mmrm(FEV1 ~ ARMCD + AVISIT + ARMCD * AVISIT + ar1(AVISIT | USUBJID), data = fev_data)
```

the design matrix is

```{r}
mx <- model.matrix(fit)
```

and the effect is

| effect |  column in design matrix |
|--------|--------------------------|
|  mu    |        1                 |
| ARMCD  |        2                 |
| AVISIT |        3,4,5             |
| ARMCD * AVISIT  |    6,7,8        |

the effect `ARMCD * AVISIT` contains `ARMCD` and `AVISIT`.

Let's calculate the `L` for `ARMCD` and `AVISIT`

### ARMCD part
```{r}
x0 <- mx[, c(1, 3, 4, 5)]
x1 <- mx[, c(2), drop = FALSE]
x2 <- mx[, c(6, 7, 8)]

l <- rep(0, 8) # initial values
m <- diag(rep(1, nrow(x0))) - x0 %*% MASS::ginv(t(x0) %*% x0) %*% t(x0)
l1_coef <- MASS::ginv(t(x1) %*% m %*% x1) %*% t(x1) %*% m %*% x1
l[c(2)] <- l1_coef

l2_coef <- MASS::ginv(t(x1) %*% m %*% x1) %*% t(x1) %*% m %*% x2
l[c(6, 7, 8)] <- l2_coef

df_md(fit, l) # very close !
```

### AVISIT part

```{r}
x0 <- mx[, c(1, 2)]
x1 <- mx[, c(3, 4, 5), drop = FALSE]
x2 <- mx[, c(6, 7, 8)]

l <- matrix(0, nrow = 3, ncol = 8) # initial values
m <- diag(rep(1, nrow(x0))) - x0 %*% MASS::ginv(t(x0) %*% x0) %*% t(x0)
l1_coef <- MASS::ginv(t(x1) %*% m %*% x1) %*% t(x1) %*% m %*% x1
l[, c(3, 4, 5)] <- l1_coef

l2_coef <- MASS::ginv(t(x1) %*% m %*% x1) %*% t(x1) %*% m %*% x2
l[, c(6, 7, 8)] <- l2_coef

df_md(fit, l) # also close
```

### ARMCD * AVISIT

```{r}
x0 <- mx[, c(1, 2, 3, 4, 5)]
x1 <- mx[, c(6, 7, 8), drop = FALSE]

l <- matrix(0, nrow = 3, ncol = 8) # initial values
m <- x0 %*% MASS::ginv(t(x0) %*% x0) %*% t(x0)
l1_coef <- MASS::ginv(t(x1) %*% m %*% x1) %*% t(x1) %*% m %*% x1
l[, c(6, 7, 8)] <- l1_coef

df_md(fit, l)
```

For the interaction, it is close to type 3 if the model is `ARMCD * AVISIT + ARMCD + AVISIT`.
But if the model is `ARMCD + AVISIT + ARMCD * AVISIT`, the result is not so close. Why?

The answer might be that the order of levels. using e3 option we can see that the contrast are different.
However in R we will not have such issue because we always use `factor` and for the interaction we also have
decided reference level!


#### Changing the order of parameter

```{r}
sas_code <- "
ods output Tests1 = Tests1 Tests2 = Tests2 Tests3 = Tests3;

PROC MIXED DATA = fev cl method=reml;
  CLASS AVISIT(ref = 'VIS4') ARMCD(ref = 'PBO') USUBJID;
  MODEL FEV1 = ARMCD*AVISIT ARMCD AVISIT / ddfm=satterthwaite htype=1,2,3 e3;
  REPEATED AVISIT / subject=USUBJID type=ar(1);
RUN;
"

res <- run_sas(sas_code)
df3_2 <- sd2df("tests3")
df2_2 <- sd2df("tests2")
```

Type 2

```{r}
df2_2
```

Type 3

```{r}
df3_2
```

Note: according to SAS guide, is a effect is not contained in any other effect, type 2,3 and 4 are equivalent.
So for `AVISIT * ARMCD` the effect should be not affected by the order of arguments. This is only because of different
contrasts.
In R we will not have this issue.

## Basic summary

With the proposed method, we have very close result to what SAS have.
And given that we do not allow singularity in fitting (those aliased cols are removed from model fitting),
the process can be further simplified to use identity matrix for the f1 cols.

Note: Zapping might be needed to remove very small numbers.

# Type III Estimable Functions

Type III is constructed assuming that we have a balanced design.

According to the SAS/STAT® 13.1 User’s Guide, type III estimable can be achieved through

For fixed effect F1:

1. For effect in the model (except F1) that contains F1, equate the coefficients in the general form of estimable function to zero
2. If necessary, equate new symbols to compound expressions in the F1 block
3. Equate all symbolic coefficients outside F1 block to a linear function of the symbols in the F1 block.

In practice, if a effect is not contained in any other tests, use its orthogonal form (identity matrix) in the corresponding submatrix, e.g. for `AVISIT * PARMAMCD`
If a effect is contained in other effects, for that part we try to equate the values.

In our case of design matrix, testing `ARMCD`, there are 4 levels in total for `AVISIT * ARMCD`.
The coefficients for the effect `ARMCD` is 1, so each value for the `AVISIT * ARMCD` is 1/4.

For each level of `AVISIT`, there are 2 levels for `AVISIT * ARMCD`.
So the coefficients for `AVISIT * ARMCD` is 1/2.

However, this only works for the simplest case that there is no restriction on the design matrix.
In general, we will not have such cases so it is fine.

Note: using Type II on perfectly balanced data, the result should be the same for Type III.

### ARMCD part
```{r}
l <- rep(0, 8) # initial values
l[2] <- 1
l[6:8] <- 1 / 4

df_md(fit, l) # very close !
```

### AVISIT part

```{r}
l <- matrix(0, nrow = 3, ncol = 8) # initial values
l[, c(3, 4, 5)] <- diag(rep(1, 3))
l[, c(6, 7, 8)] <- diag(rep(1, 3)) / 2

df_md(fit, l) # also close
```


# Prototypes

```{r}
count_levels <- function(x) {
  UseMethod("count_levels")
}
count_levels.factor <- function(x) {
  length(levels(x))
}
count_levels.character <- function(x) {
  length(unique(x))
}
count_levels.numeric <- function(x) {
  1L
}

get_l_matrix <- function(fit, var, type = c(2, 3), tol = sqrt(.Machine$double.eps)) {
  type <- match.arg(type)
  mx <- component(fit, "x_matrix")
  asg <- attr(mx, "assign")
  formula <- fit$formula_parts$model_formula
  tms <- terms(formula)
  fcts <- attr(tms, "factors")[-1, , drop = FALSE] # discard the response
  ods <- attr(tms, "order")
  idx <- which(var == colnames(fcts))
  cols <- which(asg == idx)
  vars <- row.names(fcts)
  var_types <- vapply(vars, function(x) class(fit$tmb_data$data[[x]]), FUN.VALUE = "")
  var_levels <- vapply(vars, function(x) count_levels(fit$tmb_data$data[[x]]), FUN.VALUE = 1L)
  included_vars <- vars[fcts[, idx] == 1]
  included_types <- var_types[included_vars]
  total_levels <- prod(var_levels)
  coef_rows <- length(cols)
  l_mx <- matrix(0, nrow = coef_rows, ncol = length(asg))
  l_mx[, cols] <- diag(rep(1, length(cols)))
  for (i in seq_len(ncol(fcts))) {
    x1 <- mx[, cols, drop = FALSE]
    additional_vars <- vars[which(fcts[, i] > fcts[, idx])]
    additional_types <- var_types[additional_vars]
    if (ods[i] >= ods[idx] && all(fcts[, i] >= fcts[, idx]) && all(additional_types != "numeric")) {
      if (type == 2) {
        z <- c(i, idx)
        x0 <- mx[, which(!asg %in% z), drop = FALSE]
        x2 <- mx[, which(asg == i), drop = FALSE]
        m <- diag(rep(1, nrow(x0))) - x0 %*% solve(t(x0) %*% x0) %*% t(x0)
        l_coefs <- solve(t(x1) %*% m %*% x1) %*% t(x1) %*% m %*% x2
        l_mx[, which(asg == i)] <- l_coefs
      } else if (type == 3) {
        new_cols <- which(asg == i)
        additional_levels <- vapply(additional_vars, function(x) count_levels(fit$tmb_data$data[[x]]), FUN.VALUE = 1)
        t_levels <- prod(additional_levels)
        l_mx[, which(asg == i)] <- l_mx[, which(asg == idx)] / t_levels
      }
    }
  }
  l_mx[abs(l_mx) < tol] <- 0
  l_mx
}

get_l_matrix(fit, "ARMCD", "2")
get_l_matrix(fit, "ARMCD", "3")

get_l_matrix(fit, "AVISIT", "2")
get_l_matrix(fit, "AVISIT", "3")

get_l_matrix(fit, "ARMCD:AVISIT", "2")
get_l_matrix(fit, "ARMCD:AVISIT", "3")

# call df_md on these l matrix to obtain the results.

df_md(fit, get_l_matrix(fit, "AVISIT", "3"))
df_md(fit, get_l_matrix(fit, "ARMCD:AVISIT", "3"))

df_md(fit, get_l_matrix(fit, "AVISIT", "2"))
df_md(fit, get_l_matrix(fit, "ARMCD:AVISIT", "2"))
```

## Prototypes of interfaces

To implement the anova of type II and type III, consider extending the `car::Anova` method

```{r}
Anova.mmrm_tmb <- function(fit, type) {
  vars <- colnames(attr(terms(fit$formula_parts$model_formula), "factors"))
  ret <- lapply(
    vars,
    function(x) df_md(fit, get_l_matrix(fit, x, type))
  )
  ret_df <- do.call(rbind, ret)
  row.names(ret_df) <- vars
  ret_df
}
fit <- mmrm(FEV1 ~ FEV1_BL + ARMCD + FEV1_BL * AVISIT + ar1(AVISIT | USUBJID), data = fev_data)
Anova(fit, "2")
Anova(fit, "3")

aa <- get_l_matrix(fit, "AVISIT", "3")
aa[, 7:9] <- diag(rep(0, 3))
df_md(fit, aa)
```


### How about using `CONTRAST` option in SAS?

```{r}
sas_code <- "
ods output Tests1 = Tests1 Tests2 = Tests2 Tests3 = Tests3;

PROC MIXED DATA = fev cl method=reml;
  CLASS AVISIT(ref = last) ARMCD(ref = 'PBO') USUBJID;
  MODEL FEV1 = FEV1_BL ARMCD AVISIT FEV1_BL * AVISIT / ddfm=satterthwaite htype=1,2,3 solution chisq e3;
  REPEATED AVISIT / subject=USUBJID type=ar(1);
  CONTRAST 'fev1_bl*avisit'
  FEV1_BL*AVISIT 1 0 0 -1, FEV1_BL*AVISIT 0 1 0 -1, FEV1_BL*AVISIT 0 0 1 -1;
RUN;
"

res <- run_sas(sas_code)
df3_3 <- sd2df("tests3")
df2_3 <- sd2df("tests2")
```

Apprently, if contrasts is specified as what we expected (identity matrix for FEV1_BL * AVISIT),
the inference is identical to what we have in R.
However, SAS uses a different level system (the ref is always putted to last level! and this will affect the inference)

e.g. in SAS we have the following contrast if we specify `VIS4` as the reference

| VIS1 | VIS2 | VIS3 | VIS4 |
|-|-|-|-|
| 1 | 0 | 0 | -1 |
| 0 | 1 | 0 | -1 |
| 0 | 0 | 1 | -1 |

however in our model the coefficients are estimated with VIS1 as reference.
This may also address some minor discrepencies in some results between SAS and R.

For R side, we would still use the identity matrix to test the terms.
It makes sense and should not be changed.
However, it is still possible to achieve what they have done, e.g. with the following contrast we
can have identical results

```{r}
a <- matrix(
  c(
    0, 0, -1,
    1, 0, -1,
    0, 1, -1
  ),
  nrow = 3,
  byrow = T
)
contra <- matrix(0, 3, 9)
contra[, 7:9] <- a
df_md(fit, contra)
```

Now we get very close result too!

### Additional tries (some random code of try!)

```{r}
zap_small <- function(x) {
  x[abs(x) < sqrt(.Machine$double.eps)] <- 0
  x
}

fit <- mmrm(FEV1 ~ RACE + AVISIT + RACE * AVISIT + ar1(AVISIT | USUBJID), data = fev_data)
mx <- model.matrix(fit)

x0 <- mx[, c(1, 4:6)]
x1 <- mx[, 2:3, drop = FALSE]
x2 <- mx[, c(7:12)]

l <- matrix(0, 2, 12) # initial values
m <- diag(rep(1, nrow(x0))) - x0 %*% MASS::ginv(t(x0) %*% x0) %*% t(x0)
l1_coef <- MASS::ginv(t(x1) %*% m %*% x1) %*% t(x1) %*% m %*% x1
l[, 2:3] <- l1_coef

l2_coef <- MASS::ginv(t(x1) %*% m %*% x1) %*% t(x1) %*% m %*% x2
l[, 7:12] <- l2_coef

df_md(fit, l) # very close !

fit <- mmrm(FEV1 ~ FEV1_BL * ARMCD + ARMCD + FEV1_BL + AVISIT + AVISIT * ARMCD + ar1(AVISIT | USUBJID), data = fev_data)
emmeans::joint_tests(fit)
a <- rep(0, length(coef(fit)))
a[2] <- 1
a[7] <- 0.5
df_md(fit, a)

get_l_matrix(fit, "FEV1_BL", "2")
```
