---
title: "Heterogeneous Compound Symmetry Estimation"
author: "Philippe Boileau"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, echo=FALSE}
## load required libraries
library(here)
library(rtables)
library(ggplot2)

## set global knitr options
knitr::opts_chunk$set(
  echo = FALSE, fig.width = 10, fig.height = 6, comment = ""
)
```
```{css, echo=FALSE}
.reveal .r code {
    white-space: pre;
}
```

# Description

It's been noted that models fit using `mmrm`, `glmmTMB` and `nlme` differ from
those produced by SAS's `PROC MIXED` procedure when the data-generating
processe's repeated measures' covariance matrix is assumed to be a heterogeneous
compound symmetry (CSH) matrix. This directory contains a simulation study,
relying on the [`simChef`](https://github.com/Yu-Group/simChef) simulation
framework, investigating whether this issue is due to the repeated measures'
covariance matrix estimation procedure implemented in SAS.

## Simulation Details

Two data-generating processes mimicking 1:1 randomized control trials with 10
repeated measures are considered: one in which the true covariance matrix is
homoscedastic CSH, and one in which it is heteroscedastic CSH. No other
covariates are considered. The former's variances are one at all time points.
The latter's variances increase from 0.5 at time 1 to 2.75 at time 10. The
outcome data is Gaussian conditional on treatment in both settings. One hundred
replicates are produced at sample sizes of 250, 500 and 1000 for each
data-generating process.

## Methods

Mixed models for repeated measures are fit with the `mmrm`, `glmmTMB` and `nlme`
packages' `mmrm()`, `glmmTMB()` and `gsl()` functions respectively. SAS's `PROC
MIXED` procedure is fit using the `sasr` package's SAS-R interface. All models
are fit by specifying a sole fixed effect for the treatment indicator and a
compound symmetry matrix for the repeated measures' covariance matrix.

## Benchmark Summary

Empirical risks based on the Frobenius and spectral norms are computed for each
estimator. Recall that the Frobenius norm is analogous to the squared error:
it's the square root of the sum of squared matrix elements. The spectral norm is
the square root of the largest absolute eigenvalue. The empirical mean squared
errors and biases are also computed for each of the CSH covariance matrix
parameters.

## Navigating the Directory

The main R script is `R/meal.R`. This file assembles the simulation study
components using the `simChef` framework and performs the experiment. The
data-generating function is in `R/dgp-functions.R`, and the `mmrm`,
`glmmTMB`,`nlme` and `PROC MIXED` wrapper functions are in
`R/method-functions.R`. Functions for evaluating estimator performance and
tabulating or plotting these results are contained in `R/eval-functions.R` and
`R/visualizer-functions.R` respectively. `slurm/` contains a bash script for
running this simulation on a SLURM HPC. The logs produced by running this bash
script are saved in `logs/`. Simulation study results are stored in `results/`.
Finally, the configuration file for
[`sasr`](https://github.com/insightsengineering/sasr), required for benchmarking
`PROC MIXED`, must be named `sascfg_personal.py`.


# Results

Simulation study results are stored in
`simulations/csh-benchmarks/results/hom_rct-het_rct/Varying
num_part-num_part/viz_results.rds`. 

```{r load-results}
## load the results
results <- readRDS(here(
  "simulations/csh-benchmarks/results/hom_rct-het_rct/Varying num_part-num_part/viz_results.rds")
)
```

## Frobenius and Spectral Risks

```{r risk-tibble}
results$risk_tibble
```

## Frobenius and Spectral Loss Distributions

```{r loss-dist-plot}
results$loss_dist_plot
```

## CSH Estimator Bias

```{r bias-plot}
results$bias_plot
```

## CSH Estimator MSE

```{r mse-plot}
results$mse_plot
```


# Discussion

The covariance matrix estimates produced by `glmmTMB`, `mmrm`, `nlme` and `PROC
MIXED` are virtually identical with respect to their Frobenius and spectral
risks, as well as in terms of the CSH covariance matrix parameter estimates. The
estimates output by the `glmmTMB` procedure are, in general, marginally more biased 
than the other procedures. This bias is neglible at larger sample sizes.

The results suggest that the discrepencies previously observed between mixed
models for repeated measures fit with R and SAS cannot be explained by errors in
the repeated measures' covariance matrix estimation.


# Session Information

```{r session-information}
sessionInfo()
```
