---
title: "Benchmarking Heterogeneous Compound Symmetry Covariance Matrix Estimation"
author: "Philippe Boileau and Daniel Sabanes Bove"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, echo=FALSE}
## load required libraries
library(here)
library(rtables)
library(ggplot2)

## set global knitr options
knitr::opts_chunk$set(
  echo = FALSE, fig.width = 10, fig.height = 6, comment = ""
)
```
```{css, echo=FALSE}
.reveal .r code {
    white-space: pre;
}
```

# Description

It's been noted that models fit using `mmrm`, `glmmTMB` and `nlme` differ from
those produced by SAS's `PROC MIXED` procedure when the data-generating
processe's repeated measures' covariance matrix is assumed to be a heterogeneous
compound symmetry (CSH) matrix. This directory contains a simulation study,
relying on the [`simChef`](https://github.com/Yu-Group/simChef) simulation
framework, investigating whether this issue is due to the repeated measures'
covariance matrix estimation procedure implemented in SAS.

## Simulation Details

Two data-generating processes mimicking 1:1 randomized control trials with 10
repeated measures are considered: one in which the true covariance matrix is
homoscedastic CSH, and one in which it is heteroscedastic CSH. No other
covariates are considered. The former's variances are one at all time points.
The latter's variances increase from 0.5 at time 1 to 2.75 at time 10. The
outcome data is Gaussian conditional on treatment in both settings. One hundred
replicates are produced at sample sizes of 250, 500 and 1000 for each
data-generating process.

## Methods

Mixed models for repeated measures are fit with the `mmrm`, `glmmTMB` and `nlme`
packages' `mmrm()`, `glmmTMB()` and `gsl()` functions respectively. SAS's `PROC
MIXED` procedure is fit using the `sasr` package's SAS-R interface. All models
are fit by specifying a sole fixed effect for the treatment indicator and a
compound symmetry matrix for the repeated measures' covariance matrix.

## Benchmark Summary

Empirical risks based on the Frobenius and spectral norms are computed for each
estimator. Recall that the Frobenius norm is analogous to the squared error:
it's the square root of the sum of squared matrix elements. The spectral norm is
the square root of the largest absolute eigenvalue. The empirical mean squared
errors and biases are also computed for each of the CSH covariance matrix
parameters.

## Navigating the Directory

The main R script is `R/meal.R`. This file assembles the simulation study
components using the `simChef` framework and performs the experiment. The
data-generating function is in `R/dgp-functions.R`, and the `mmrm`,
`glmmTMB`,`nlme` and `PROC MIXED` wrapper functions are in
`R/method-functions.R`. Functions for evaluating estimator performance and
tabulating or plotting these results are contained in `R/eval-functions.R` and
`R/visualizer-functions.R` respectively. `slurm/` contains a bash script for
running this simulation on a SLURM HPC. The logs produced by running this bash
script are saved in `logs/`. Simulation study results are stored in `results/`.
Finally, the configuration file for
[`sasr`](https://github.com/insightsengineering/sasr), required for benchmarking
`PROC MIXED`, must be named `sascfg_personal.py`.


# Results

Simulation study results are stored in
`simulations/csh-benchmarks/results/hom_rct-het_rct/Varying
num_part-num_part/viz_results.rds`. 

```{r load-results}
## load the results
results <- readRDS(here(
  "simulations/csh-benchmarks/results/hom_rct-het_rct/Varying num_part-num_part/viz_results.rds")
)
```

## Frobenius and Spectral Risks

```{r risk-tibble}
results$risk_tibble
```

## Frobenius and Spectral Loss Distributions

```{r loss-dist-plot}
results$loss_dist_plot
```

## CSH Estimator Bias

```{r bias-plot}
results$bias_plot
```

## CSH Estimator MSE

```{r mse-plot}
results$mse_plot
```


# Discussion

The covariance matrix estimates produced by `glmmTMB`, `mmrm`, `nlme` and `PROC
MIXED` are virtually identical with respect to their Frobenius and spectral
risks, as well as in terms of the CSH covariance matrix parameter estimates. The
estimates output by the `glmmTMB` procedure are, in general, marginally more biased 
than the other procedures. This bias is neglible at larger sample sizes.

The results suggest that the discrepencies previously observed between mixed
models for repeated measures fit with R and SAS cannot be explained by errors in
the repeated measures' covariance matrix estimation. They might instead be due
to a difficult to diagnosis re-labelling issue performed by SAS "under the
hood". See the example below.

## Time Point Re-Labelling Issue

Consider the following data-generating function:

```{r dgp, echo=TRUE}
#' A Simple RCT Data-Generating Function
#'
#' @description This function simulates a randomized control trial with
#'   participants equally split across the control and treatment arms.
#'   Participants' repeated measures are assumed to be normally distributed with
#'   heterogeneous compound symmetry covariance matrix. Participants' outcomes
#'   are collected at each time point.
#'
#' @param num_part A numeric indicating the total number of participants.
#' @param num_rep_meas A numeric indicating the number of repeated measures.
#' @param fixed_int A numeric corresponding to the fixed intercept effect.
#' @param fixed_trt A numeric corresponding to the fixed treatment effect.
#' @param outcome_vars A numeric vector with length equal to num_rep_meas. It is
#'   the vector of the repeated measures' variances.
#' @param outcome_cor A numeric corresponding to the correlation parameter in
#'   the heterogeneous compound symmetry covariance matrix.
#' @param pad_zeroes A logical indicating whether to pad the time points with a
#'   leading zero. If set to FALSE, the default, then the time variable is an
#'   integer. If set to true, the time variable is transformed into an ordered
#'   factor.
#'
#' @return A list containing the simulated participants' IDs, time point
#'   indicators, outcomes, treatment assignments and baseline covariates.
rct_dgp_fun <- function(
  num_part = 1000,
  num_rep_meas = 10,
  fixed_int = 1,
  fixed_trt = 1,
  outcome_vars = seq_len(num_rep_meas),
  outcome_cor = 0.5,
  pad_zeroes = FALSE
) {

  ## form a balanced data.frame
  cov_df <- data.frame(
    participant = seq_len(num_part),
    trt = c(rep(0, round(num_part / 2)), rep(1, num_part - round(num_part / 2)))
  )
  time_point_df <- expand.grid(
    time = seq_len(num_rep_meas),
    participant = seq_len(num_part)
  )
  df <- cov_df %>% dplyr::left_join(time_point_df, by = "participant")

  ## produce the model matrix for the fixed effects
  fixed_model_mat <- model.matrix(~ trt, data = df)

  ## define the repeated measures correlation structure, assuming
  ## heterogeneous compound symmetry
  csh_mat <- tcrossprod(sqrt(outcome_vars), sqrt(outcome_vars)) * outcome_cor
  diag(csh_mat) <- outcome_vars

  ## generate the outcomes
  beta <- c(fixed_int, fixed_trt)
  df$y <- fixed_model_mat %*% beta +
    as.vector(t(MASS::mvrnorm(num_part, rep(0, num_rep_meas), csh_mat)))

  ## time point variable transformation
  if (pad_zeroes) {
    df$time <- stringr::str_pad(df$time, width = 2, side = "left", pad = "0")
  }

  ## return the generated data
  return(list(
    participant = as.factor(df$participant),
    time = as.factor(df$time),
    y = df$y,
    trt = df$trt
  ))
}
```

We apply the `PROC MIXED` wrapper function to datasets that are identical in all
but their `time` variables. This function returns parameter estimates for the
CSH covariance matrix. In the first dataset, `time` is a factor with levels `1`,
`2`, ..., `10`. The second dataset's `time` variable is a factor with levels
`01`, `02`, ..., `10`.

```{r run-sasr, echo=TRUE}
## generate datasets with and without time points padded by zeroes
set.seed(123)
rct_data_not_padded <- rct_dgp_fun()
rct_data_not_padded$time %>% head(10)
set.seed(123)
rct_data_padded <- rct_dgp_fun(pad_zeroes = TRUE)
rct_data_padded$time %>% head(10)

## load the sasr wrapper function
source(here("simulations/csh-benchmarks/R/method-functions.R"))
proc_mixed_fun

## fit the data to the non-paded time point data
proc_mixed_fun(
  participant = rct_data_not_padded$participant,
  time = rct_data_not_padded$time,
  y = rct_data_not_padded$y,
  trt = rct_data_not_padded$trt
)

## fit the data to the paded time point data
proc_mixed_fun(
  participant = rct_data_padded$participant,
  time = rct_data_padded$time,
  y = rct_data_padded$y,
  trt = rct_data_padded$trt
)
```

When the `time` isn't padded by zeroes, we find that second and tenth time
points' variances are swapped. The `time` data appears to be loaded as a string
into the SAS session and subsequently sorted. That is, `"1", "2", ..., "10"` is
reordered as `"1", "10", "2", ..., "9"`. That's not the entire story, though, as
the names of the covariance parameters returned by applying the wrapper function
to the dataset with padded `time` aren't padded. In any case, this issue should
be taken into account when comparing `mmrm()` and `PROC MIXED` ouputs.


# Session Information

```{r session-information}
sessionInfo()
```
